\documentclass{article}

% Core packages
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{mathtools}
\usepackage{bm}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{subcaption}
\usepackage{enumitem}
\usepackage[margin=1in]{geometry}
\usepackage{natbib}
\usepackage{amsthm}

\newtheorem{proposition}{Proposition}

% Custom commands
\newcommand{\vx}{\bm{x}}
\newcommand{\vo}{\bm{o}}
\newcommand{\vz}{\bm{z}}
\newcommand{\va}{\bm{a}}
\newcommand{\vb}{\bm{b}}
\newcommand{\vep}{\bm{\epsilon}}
\newcommand{\vh}{\bm{h}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\KL}{\mathrm{KL}}
\newcommand{\loss}{\mathcal{L}}
\DeclareMathOperator{\sigmoid}{sigmoid}
\DeclareMathOperator{\diag}{diag}

\title{Action-Conditioned Unified Latents for Robotic World Models}

\author{
  Anonymous Authors
}

\date{}

\begin{document}

\maketitle

\begin{abstract}
Learning compact, physics-grounded visual representations is a central challenge for building robotic world models that jointly predict future observations and actions.
Existing approaches either use generic pretrained video VAEs---whose latent spaces retain task-irrelevant visual details---or frozen pretrained features that discard action-relevant geometric information.
We introduce \textbf{Action-Conditioned Unified Latents (AC-UL)}, a framework that applies the Unified Latents (UL) variational objective to \emph{fine-tune} a pretrained video model for autoregressive state-action dynamics.
Starting from a pretrained video VAE and diffusion transformer, we add a learned action encoder and jointly fine-tune the system so that the dynamics prior must \emph{jointly denoise} the current action latent $\vb_t$ and the next observation latent $\vz_{t+1}$---the frame that results from executing the action---unifying world modeling and policy learning in a single generative process.
Both observations and actions are encoded into compact latent spaces and denoised with a shared flow matching objective.
By conditioning the diffusion prior on the observation-action latent history, proprioceptive state, and language goal, the variational bound incentivizes the encoder to discard unpredictable content while retaining the physical geometry that the history of interactions can predict.
In Stage~1, a pretrained video VAE and a pretrained diffusion transformer (serving as the dynamics prior) are jointly fine-tuned with LoRA under the AC-UL objective alongside newly trained action encoder/decoder components; in Stage~2, the frozen fine-tuned encoders provide a physics-grounded vocabulary on which the autoregressive diffusion transformer is further fine-tuned for joint video-action generation, enabling closed-loop planning via iterative rollouts.
We evaluate AC-UL on the DROID real-world manipulation dataset and show that a 1.3B AC-UL model can match or outperform the 14B DreamZero baseline, demonstrating that adapting the latent space for action-relevant content is more effective than scaling model parameters on generic representations.
\end{abstract}

%=====================================================================
\section{Introduction}
\label{sec:intro}
%=====================================================================

Building world models---internal simulators that predict how actions change the environment---is a long-standing goal in robotics and reinforcement learning~\citep{ha2018world,hafner2023dreamerv3}.
A world model that operates in a learned latent space can be orders of magnitude cheaper to roll out than a pixel-level simulator, enabling model-predictive control, planning, and data-efficient policy learning.

Recent progress in autoregressive video-action models~\citep{dreamzero2025} has shown that jointly generating future video frames and robot actions from a shared backbone can achieve remarkable generalization.
Meanwhile, latent diffusion models~\citep{rombach2022high} and vision-language-action (VLA) architectures~\citep{brohan2023rt2,kim2024openvla} have produced powerful visual representations.
However, two fundamental problems remain:

\textbf{(1) Generic VAE latent spaces} (e.g., DreamZero~\citep{dreamzero2025} uses a Wan2.1 VAE~\citep{wan2025}) retain textures, lighting, and background clutter that are irrelevant to action dynamics.
The VAE was trained for generic video reconstruction, not for robotic manipulation, so its latent space confounds physical geometry with appearance.

\textbf{(2) Frozen-feature approaches} (e.g., DINO-WM~\citep{zhou2024dinowm}) use pretrained encoders (DINOv2) that were not trained to understand action effects.
While these features transfer surprisingly well, they are agnostic to the causal structure of manipulation.

We propose a third path: \textbf{fine-tuning a pretrained video model's latent space to encode only the information that the history of interactions can predict}.
Our approach applies the Unified Latents (UL) framework~\citep{heek2025ul} to \emph{adapt} a pretrained video VAE and diffusion transformer for joint observation-action dynamics.
The key insight is that by adding a learned action encoder and conditioning the diffusion prior on the latent history, proprioceptive state, and a language instruction, any encoder information that cannot be predicted from this context will spike the variational loss.
Both the action latent $\vb_t$ and the observation latent $\vz_{t+1}$ are denoised with the same flow matching objective, ensuring a unified generative process.
The encoder is thus incentivized to adapt its latent space---starting from powerful pretrained features---to discard unpredictable content and retain only the physical geometry that the robot's interaction history and task goal can explain.

This formulation follows the autoregressive structure of DreamZero~\citep{dreamzero2025}, but applies the UL variational objective to fine-tune the latent space for action-relevant content, rather than using a frozen generic VAE.

Our contributions are:
\begin{enumerate}[leftmargin=*,itemsep=2pt]
  \item We derive the \textbf{Action-Conditioned Unified Latent (AC-UL)} objective, extending the UL variational bound to history-conditioned \emph{joint} observation-action dynamics with learned action encoding and a unified flow matching objective for both modalities (\Cref{sec:method}).
  \item We provide a complete \textbf{two-stage training algorithm}: Stage~1 fine-tunes a pretrained video VAE and a pretrained diffusion transformer (as dynamics prior) with LoRA under the AC-UL objective alongside learned action encoder/decoder components; Stage~2 further fine-tunes the same pretrained diffusion transformer on frozen AC-UL latents for autoregressive joint video-action generation (\Cref{sec:stage2}).
  \item We demonstrate on the \textbf{DROID} real-world manipulation dataset that a \textbf{1.3B AC-UL model matches or outperforms a 14B DreamZero baseline} across closed-loop planning, imitation learning, and video-action prediction, showing that better latent representations outweigh raw model scale (\Cref{sec:experiments}).
\end{enumerate}

%=====================================================================
\section{Background}
\label{sec:background}
%=====================================================================

\subsection{Variational Autoencoders and the ELBO}

Given data $\vx$ and latent variable $\vz_0$, the Evidence Lower Bound (ELBO) on the log-likelihood is:
\begin{equation}
  -\log p_\theta(\vx) \leq \E_{\vz_0 \sim p_\theta(\vz_0|\vx)}\!\Big[\underbrace{-\log p_\theta(\vx|\vz_0)}_{\text{decoder}}\Big] + \KL\!\Big[\underbrace{p_\theta(\vz_0|\vx)}_{\text{encoder}} \,\Big\|\, \underbrace{p_\theta(\vz_0)}_{\text{prior}}\Big] = \loss(\vx).
  \label{eq:elbo}
\end{equation}
In the Unified Latents framework, both the decoder $p_\theta(\vx|\vz_0)$ and the prior $p_\theta(\vz_0)$ are parameterized as diffusion models.

\subsection{Diffusion Models and the KL Bound}

Consider a destruction process $\vx_t = \alpha(t)\,\vx + \sigma(t)\,\vep$ with $\vep \sim \mathcal{N}(\bm{0},\bm{I})$, where the noise level is characterized by the log signal-to-noise ratio $\lambda(t) = \log(\alpha_t^2/\sigma_t^2)$ with the variance-preserving constraint $\alpha_t^2 + \sigma_t^2 = 1$.
The KL divergence between the data-conditioned diffusion path and the learned reverse process can be bounded as~\citep{ho2020denoising,kingma2021variational}:
\begin{equation}
  \KL\!\Big[p(\vx_0|\vx) \,\Big\|\, p_\theta(\vx_0)\Big]
  \leq \E_{t\sim\mathcal{U}(0,1)}\!\left[-\frac{d\lambda(t)}{dt}\,\frac{\exp\lambda(t)}{2}\,w(\lambda_t)\,\|\vx - \hat{\vx}(\vx_t,\theta)\|^2\right]
  + \KL\!\Big[p(\vx_1|\vx) \,\Big\|\, \mathcal{N}(\bm{0},\bm{I})\Big],
  \label{eq:diff_kl}
\end{equation}
where $w(\lambda_t) = 1$ yields a tight bound, and $w(\lambda_t) = \sigmoid(\lambda_t - b)$ provides a perceptually motivated re-weighting.

\subsection{Unified Latents (UL)}
\label{sec:ul_background}

UL~\citep{heek2025ul} uses a deterministic encoder $\vz_{\text{clean}} = E_\theta(\vx)$ with fixed Gaussian noise added at the minimum noise level:
\begin{equation}
  \vz_0 = \alpha_z(0)\,\vz_{\text{clean}} + \sigma_z(0)\,\vep_z, \quad
  \alpha_z(0) \approx 1.0, \quad
  \sigma_z(0) \approx 0.08.
  \label{eq:ul_noise}
\end{equation}

\textbf{Prior loss (Equation~3 of UL).}
The KL penalty between the encoder distribution and the diffusion prior is:
\begin{equation}
  \KL\!\Big[p(\vz_0|\vx) \,\Big\|\, p_\theta(\vz_0)\Big]
  \leq \E_{\tau}\!\left[-\frac{d\lambda_z(\tau)}{d\tau}\,\frac{\exp\lambda_z(\tau)}{2}\,w(\lambda_z(\tau))\,\|\vz_{\text{clean}} - \hat{\vz}(\vz_\tau, \theta)\|^2\right]
  + \KL\!\Big[p(\vz_1|\vx) \,\Big\|\, \mathcal{N}(\bm{0},\bm{I})\Big],
  \label{eq:ul_prior}
\end{equation}
where $w(\lambda_z(\tau)) = 1$ for a tight bitrate bound. This bound holds for any noise schedule $\vz_\tau = \alpha(\tau)\vz_{\text{clean}} + \sigma(\tau)\vep$, and we will refer to the loss-weighting function $w_{\KL}(\tau) \triangleq -\frac{d\lambda(\tau)}{d\tau}\frac{\exp\lambda(\tau)}{2}$ as the \emph{KL-optimal weighting}.

\textbf{Decoder loss (Equation~4 of UL).}
The reconstruction term uses a diffusion decoder conditioned on $\vz_0$:
\begin{equation}
  -\log p_\theta(\vx|\vz_0) \leq \E_{\tau\sim\mathcal{U}(0,1)}\!\left[\frac{d\lambda_x(\tau)}{d\tau}\,\frac{\exp\lambda_x(\tau)}{2}\,w_x(\lambda_x(\tau))\,\|\vx - \hat{\vx}(\vx_\tau, \vz_0, \theta)\|^2\right],
  \label{eq:ul_decoder}
\end{equation}
with sigmoid weighting $w_x(\lambda_x(\tau)) = \sigmoid(\lambda_x(\tau) - b)$ and a loss factor $c_{\text{lf}} \in [1.3, 1.7]$ that up-weighs the decoder to prevent posterior collapse.
In our method, we depart from UL by using a pretrained deterministic decoder instead of a diffusion decoder (see \Cref{sec:decoder}).

\subsection{Flow Matching}
\label{sec:flow_matching_background}

Flow matching~\citep{lipman2023flow,liu2023flow} provides an alternative to DDPM for training continuous generative models. Instead of learning a score function under a variance-preserving noise schedule, flow matching learns a \emph{velocity field} $v_\theta$ that transports samples along a linear interpolation path:
\begin{equation}
  \vx_\tau = (1-\tau)\,\vx + \tau\,\vep, \quad \vep \sim \mathcal{N}(\bm{0}, \bm{I}), \quad \tau \in [0,1],
  \label{eq:fm_interpolation}
\end{equation}
where $\tau=0$ corresponds to clean data and $\tau=1$ to pure noise. The conditional velocity field target is $\vx - \vep$ (the direction from noise to data), and the training objective is:
\begin{equation}
  \loss_{\text{FM}} = \E_{\tau, \vep}\!\left[\left\|v_\theta(\vx_\tau, \tau) - (\vx - \vep)\right\|^2\right].
  \label{eq:fm_loss}
\end{equation}

\textbf{Relationship to DDPM.}
Both DDPM and flow matching can be expressed as weighted denoising objectives. The velocity prediction loss is equivalent to a \emph{clean-sample prediction} loss with $\tau$-dependent weighting (see \Cref{sec:fm_approx} for the full derivation):
\begin{equation}
  \left\|v_\theta - (\vx - \vep)\right\|^2 = \frac{1}{\tau^2}\left\|\vx - \hat{\vx}_\theta\right\|^2, \quad \text{where } \hat{\vx}_\theta \triangleq \vx_\tau + \tau\, v_\theta.
  \label{eq:fm_ddpm_equiv}
\end{equation}
This means the flow matching loss is a reweighted version of the DDPM denoising loss, with implicit weighting $w_{\text{FM}}(\tau) = 1/\tau^2$ instead of the KL-optimal weighting $w_{\KL}(\tau)$.

\subsection{Autoregressive Video-Action Models}
\label{sec:dreamzero_background}

DreamZero~\citep{dreamzero2025} introduced an autoregressive flow-matching transformer that jointly generates video frames and robot actions.
The model uses a frozen Wan2.1 VAE~\citep{wan2025} to encode video into latent tokens and a CausalWanModel (14B parameters, fine-tuned with LoRA from a pretrained image-to-video model) as the diffusion backbone. Wan2.1 is also available at 1.3B scale, which we adopt for AC-UL.
Given a history of VAE-encoded observations and past actions, the model jointly generates the current action and the next video latent using blockwise causal attention with flow matching.
Actions are processed by embodiment-specific MLPs within the diffusion model and denoised alongside video tokens.

Concretely, the model takes observations $\vo_{t-H+1:t}$, past actions $\va_{t-H:t-1}$, proprioceptive state $\bm{s}_t$, and a language instruction $\bm{l}$, and jointly generates:
\begin{equation}
  p_\theta(\va_t, \vo_{t+1} \mid \vo_{t-H+1:t}, \va_{t-H:t-1}, \bm{s}_t, \bm{l}).
  \label{eq:dreamzero}
\end{equation}
The generation is autoregressive across time steps: at step $t{-}1$ the model generated $(\va_{t-1}, \vo_t)$, and now at step $t$ it generates $(\va_t, \vo_{t+1})$, always producing the action that causes the next frame.
During closed-loop execution, predicted frames are replaced with ground-truth observations after each action, preventing error accumulation.

We adopt this autoregressive structure but apply the UL variational objective to \emph{fine-tune} the VAE's latent space for action-relevant content, rather than using it as-is.

%=====================================================================
\section{Method: Action-Conditioned Unified Latents}
\label{sec:method}
%=====================================================================

We apply the UL framework to \emph{fine-tune} a pretrained video model for \emph{history-conditioned joint action-observation dynamics}.
The central idea is to (1)~initialize the observation encoder $E_\theta$ and decoder $D_\theta$ from a pretrained video VAE (e.g., Wan2.1~\citep{wan2025}), (2)~add a learned action encoder $A_\theta$ and action decoder $D_\theta^a$, and (3)~replace the unconditional prior $p_\theta(\vz_0)$ with a \emph{dynamics prior} $p_\theta(\vb_t, \vz_{t+1,0} \mid \mathcal{H}_t^z)$ that must jointly denoise the current action latent and the resulting future observation latent using a unified flow matching objective, conditioned on the interaction history, proprioceptive state, and a language instruction.
By fine-tuning the pretrained VAE with the AC-UL loss, we leverage powerful visual features learned from large-scale video data while adapting the latent space to be action-grounded.

\subsection{Problem Setup and Variables}
\label{sec:variables}

We consider a robotic agent that observes image frames (potentially from multiple camera views), receives proprioceptive state, a language instruction, and executes actions in an autoregressive loop. We define the \emph{interaction history} at time $t$:
\begin{equation}
  \mathcal{H}_t = \big(\bm{l},\; \bm{s}_t,\; \vo_{t-H+1}, \va_{t-H}, \vo_{t-H+2}, \va_{t-H+1}, \ldots, \vo_{t-1}, \va_{t-1}, \vo_t\big),
  \label{eq:history}
\end{equation}
where:
\begin{itemize}[leftmargin=*,itemsep=2pt]
  \item $\bm{l}$: a natural language instruction specifying the task (e.g., ``pick up the red block'').
  \item $\bm{s}_t \in \R^{d_s}$: proprioceptive state at time $t$ (e.g., joint positions + gripper state).
  \item $\vo_k \in \R^{H_{\text{img}} \times W_{\text{img}} \times 3}$: RGB observation at time step $k$ (may be a multi-view composite image following DreamZero).
  \item $\va_k \in \R^{d_a}$: robot action executed at step $k$ (e.g., joint position deltas + gripper command).
  \item $H$: history window length.
\end{itemize}

Note the autoregressive structure: the history contains observations $\vo_{t-H+1:t}$ and actions $\va_{t-H:t-1}$. At the previous time step, the model generated $(\va_{t-1}, \vo_t)$ as a pair---the action $\va_{t-1}$ \emph{caused} the transition to $\vo_t$. Now at time $t$, the model's task is to jointly generate:
\begin{equation}
  (\va_t, \vo_{t+1}) \sim p_\theta(\va_t, \vo_{t+1} \mid \mathcal{H}_t),
  \label{eq:joint_gen}
\end{equation}
where $\va_t$ is the action to execute now, and $\vo_{t+1}$ is the observation that \emph{results from executing} $\va_t$. The action $\va_t$ and the next frame $\vo_{t+1}$ are thus causally linked: the model must generate an action consistent with the predicted future.

\subsection{Latent Encoding and the Forward Process}
\label{sec:encoding}

\textbf{Observation encoder.}
The observation encoder $E_\theta$ is initialized from a pretrained video VAE encoder (e.g., Wan2.1~\citep{wan2025}) and compresses each observation into a deterministic latent:
\begin{equation}
  \vz_{k,\text{clean}} = E_\theta(\vo_k) \in \R^{h \times w \times c}, \quad k = t-H+1, \ldots, t, t+1,
  \label{eq:encode}
\end{equation}
with spatial dimensions $h \times w$ downsampled from $H_{\text{img}} \times W_{\text{img}}$ and $c$ latent channels.
By initializing from a pretrained VAE, the encoder starts with strong visual features and is fine-tuned by the AC-UL objective to adapt the latent space for action-relevant content.

\textbf{Action encoder.}
The action encoder $A_\theta$ maps each raw action into a deterministic action latent:
\begin{equation}
  \vb_{k,\text{clean}} = A_\theta(\va_k) \in \R^{d_b}, \quad k = t-H, \ldots, t-1, t,
  \label{eq:action_encode}
\end{equation}
where $d_b$ is the action latent dimension. This encoding allows the action to be treated symmetrically with observation latents in the flow matching process.

The latent history is constructed from encoded observations, encoded actions, and proprioceptive state:
\begin{equation}
  \mathcal{H}_t^z = \big(\bm{l},\; \bm{s}_t,\; \vz_{t-H+1,\text{clean}}, \vb_{t-H,\text{clean}}, \vz_{t-H+2,\text{clean}}, \vb_{t-H+1,\text{clean}}, \ldots, \vb_{t-1,\text{clean}}, \vz_{t,\text{clean}}\big).
  \label{eq:latent_history}
\end{equation}

\textbf{Forward process (flow matching interpolation).}
Since the pretrained Wan2.1 model uses flow matching, we adopt the same linear interpolation schedule for the forward process. For any time $\tau \in [0,1]$:
\begin{align}
  \vz_{t+1,\tau} &= (1-\tau)\,\vz_{t+1,\text{clean}} + \tau\,\vep_z, \quad \vep_z \sim \mathcal{N}(\bm{0}, \bm{I}),
  \label{eq:forward_latent} \\
  \vb_{t,\tau} &= (1-\tau)\,\vb_{t,\text{clean}} + \tau\,\vep_b, \quad \vep_b \sim \mathcal{N}(\bm{0}, \bm{I}),
  \label{eq:forward_action}
\end{align}
where $\tau = 0$ corresponds to the clean signal and $\tau = 1$ to pure noise.
The same time $\tau$ is shared between the observation and action latents, enabling joint denoising.

\textbf{Relationship to the ELBO.}
In the original UL framework (DDPM), a stochastic encoder is obtained by adding fixed Gaussian noise at a minimum noise level $\sigma(0) > 0$, yielding a proper encoder distribution $q(\vz_0 \mid \vo) = \mathcal{N}(\alpha(0)\,\vz_{\text{clean}},\; \sigma(0)^2 \bm{I})$.
The KL divergence $\KL[q(\vz_0 \mid \vo) \| p_\theta(\vz_0)]$ is then finite and upper-bounded by integrating the denoising loss over all noise levels $\tau$.
In flow matching, the interpolation gives $\vz_0 = \vz_{\text{clean}}$ at $\tau = 0$---a Dirac delta, not a Gaussian---so the formal KL (and hence the ELBO) is not directly available.
However, this does not affect the practical training objective: the FM velocity loss penalizes the clean-sample prediction error $\|\vz_{\text{clean}} - \hat{\vz}_\theta\|^2$ at every $\tau > 0$, and this error increases whenever the encoder encodes content that the prior cannot predict from the history.
The information bottleneck mechanism---discarding unpredictable content while retaining history-predictable content (\Cref{eq:info_decomp})---is therefore preserved as an informal but rigorous directional argument, even though the FM loss is not a formal upper bound on the KL.
We discuss this relaxation and its implications in \Cref{sec:theory}.

\subsection{History-Conditioned Joint Dynamics Prior}
\label{sec:ac_prior}

This is the central contribution. In standard UL, the prior $\hat{\vz}(\vz_t, \theta)$ unconditionally denoises the latent.
We replace this with a \emph{joint dynamics prior} that conditions on the full interaction history $\mathcal{H}_t^z$ (including language instruction) and simultaneously predicts the velocity field for the current action latent and the resulting future observation latent:
\begin{equation}
  \big(v_\theta^b,\; v_\theta^z\big) = \hat{f}_\theta^{\text{dyn}}\!\left(\vb_{t,\tau},\; \vz_{t+1,\tau},\; \mathcal{H}_t^z,\; \tau\right).
  \label{eq:dynamics_prior_pred}
\end{equation}

The prior jointly predicts velocities for:
\begin{enumerate}[leftmargin=*,itemsep=2pt]
  \item \textbf{Current action latent:} $v_\theta^b \approx \vb_{t,\text{clean}} - \vep_b$ --- the velocity from noise to the encoded action, given the history and language goal.
  \item \textbf{Future observation latent:} $v_\theta^z \approx \vz_{t+1,\text{clean}} - \vep_z$ --- the velocity from noise to the observation that results from executing the action.
\end{enumerate}
Both are generated from noised inputs using the same flow matching process. The clean-sample predictions are recovered as $\hat{\vb}_t = \vb_{t,\tau} + \tau\, v_\theta^b$ and $\hat{\vz}_{t+1} = \vz_{t+1,\tau} + \tau\, v_\theta^z$. The action latent and future observation latent are causally linked: the decoded action $D_\theta^a(\hat{\vb}_t)$ is the action that \emph{causes} the transition $\vo_t \to \vo_{t+1}$. The model must predict both velocities consistently.

\subsubsection{Derivation: From DDPM KL Bound to Flow Matching Prior}
\label{sec:fm_derivation}

We first recall the original UL derivation, then show how the flow matching prior loss approximates it.

\textbf{Step 1: The DDPM KL bound (original UL).}
The ideal prior loss in the UL framework upper-bounds the KL divergence between the encoder and prior distributions. For the joint latent $(\vz_{t+1,0}, \vb_{t,0})$, the joint KL decomposes additively (by independence of the deterministic encoders):
\begin{align}
  &\KL\!\Big[p(\vz_{t+1,0}, \vb_{t,0} \mid \vo_{t+1}, \va_t) \;\Big\|\; p_\theta(\vz_{t+1,0}, \vb_{t,0} \mid \mathcal{H}_t^z)\Big] \nonumber \\
  &\leq \underbrace{\E_\tau\!\left[w_{\KL}(\tau)\,\|\vz_{t+1,\text{clean}} - \hat{\vz}_\theta\|^2\right]}_{\loss_z^{\text{KL}}} + \gamma \cdot \underbrace{\E_\tau\!\left[w_{\KL}(\tau)\,\|\vb_{t,\text{clean}} - \hat{\vb}_\theta\|^2\right]}_{\loss_b^{\text{KL}}} + C_{\KL},
  \label{eq:ddpm_kl_bound}
\end{align}
where $w_{\KL}(\tau) = -\frac{d\lambda(\tau)}{d\tau}\frac{\exp\lambda(\tau)}{2}$ is the KL-optimal weighting (\Cref{eq:ul_prior}), and $C_{\KL}$ collects the boundary KL terms. This is a tight bound when $w = w_{\KL}$.

\textbf{Step 2: Flow matching as a reweighted denoising loss.}
The pretrained Wan2.1 model uses flow matching with linear interpolation $\vz_\tau = (1-\tau)\vz_{\text{clean}} + \tau\vep$. As shown in \Cref{sec:fm_approx}, the velocity prediction loss is equivalent to a clean-sample prediction loss with implicit weighting $1/\tau^2$:
\begin{equation}
  \|v_\theta - (\vz_{\text{clean}} - \vep)\|^2 = \frac{1}{\tau^2}\,\|\vz_{\text{clean}} - \hat{\vz}_\theta\|^2,
  \quad\text{where } \hat{\vz}_\theta = \vz_\tau + \tau\, v_\theta.
  \label{eq:fm_reweight}
\end{equation}

Therefore, both the DDPM and flow matching prior losses have the general form $\E_\tau[w(\tau)\,\|\vz_{\text{clean}} - \hat{\vz}_\theta\|^2]$, differing only in the weighting function:
\begin{equation}
  w_{\KL}(\tau) = -\tfrac{d\lambda}{d\tau}\tfrac{e^\lambda}{2} \quad\text{(tight KL bound)}, \qquad
  w_{\text{FM}}(\tau) = \tfrac{1}{\tau^2} \quad\text{(flow matching)}.
  \label{eq:weight_comparison}
\end{equation}

\textbf{Step 3: The flow matching prior loss.}
We adopt the flow matching velocity prediction loss as our practical prior objective:

\textbf{Observation latent loss:}
\begin{equation}
  \loss_z^{\text{latent}}(\theta) = \E_{\tau \sim p(\tau)}\!\left[
    \left\|v_\theta^z\!\left(\vb_{t,\tau},\, \vz_{t+1,\tau},\, \mathcal{H}_t^z,\, \tau\right) - (\vz_{t+1,\text{clean}} - \vep_z)\right\|^2
  \right].
  \label{eq:ac_prior_loss}
\end{equation}

\textbf{Action latent loss:}
\begin{equation}
  \loss_b(\theta) = \E_{\tau \sim p(\tau)}\!\left[
    \left\|v_\theta^b\!\left(\vb_{t,\tau},\, \vz_{t+1,\tau},\, \mathcal{H}_t^z,\, \tau\right) - (\vb_{t,\text{clean}} - \vep_b)\right\|^2
  \right].
  \label{eq:action_latent_loss}
\end{equation}

Here $p(\tau)$ is a shifted Beta distribution (following DreamZero) that concentrates sampling on intermediate noise levels. The complete prior loss combines both:
\begin{equation}
  \loss_z^{\text{joint}}(\theta) = \loss_z^{\text{latent}}(\theta) + \gamma \cdot \loss_b(\theta),
  \label{eq:joint_prior_loss}
\end{equation}
where $\gamma$ controls the relative weight of action denoising and both losses share the same sampled time $\tau$.

\subsubsection{Why History-Conditioned Joint Denoising Forces Physics-Grounded Encoding}

The loss is driven by two velocity prediction MSE terms, which are equivalent to weighted clean-sample prediction errors (\Cref{eq:fm_reweight}): (i) the observation latent error $\|v_\theta^z - (\vz_{t+1,\text{clean}} - \vep_z)\|^2 = \frac{1}{\tau^2}\|\vz_{t+1,\text{clean}} - \hat{\vz}_{t+1}\|^2$ and (ii) the action latent error $\|v_\theta^b - (\vb_{t,\text{clean}} - \vep_b)\|^2 = \frac{1}{\tau^2}\|\vb_{t,\text{clean}} - \hat{\vb}_t\|^2$.

We categorize visual content by whether the dynamics prior can predict it from the latent history $\mathcal{H}_t^z$.

\textbf{Unpredictable content} (penalized by $\loss_z^{\text{latent}}$). If the encoder preserves information that the prior cannot predict from the history, the observation latent MSE spikes. Concrete examples in real-world manipulation (e.g., DROID):
\begin{enumerate}[leftmargin=*]
  \item \textbf{Camera auto-exposure adjustments:} Mid-trajectory brightness shifts caused by the camera's automatic gain control have no causal link to the robot's actions or the language instruction. Nothing in $\mathcal{H}_t^z$ predicts them.

  \item \textbf{Specular reflections:} Highlights on metallic surfaces shift as viewpoint or lighting changes. While partially correlated with gripper motion, these high-frequency appearance changes are difficult for the prior to model and are effectively unpredictable.

  \item \textbf{External disturbances:} Another person's hand entering the workspace, or an object being moved by someone else, is not explained by the robot's action history.
\end{enumerate}

\textbf{Predictable content} (``free'' under the prior loss). Information that the prior \emph{can} predict from $\mathcal{H}_t^z$ incurs no prior penalty, so the encoder is free to retain it:
\begin{enumerate}[leftmargin=*]
  \item \textbf{Action-caused geometry:} The gripper closing on the red block after a ``pick up red block'' instruction is predictable: the prior can reason from the language goal, the history of gripper approach, and the denoised action latent $\hat{\vb}_t$ that the next frame should show contact. Encoding gripper pose, object displacement, and contact geometry \emph{decreases} both MSE terms.

  \item \textbf{Object state changes:} A drawer sliding open after a pull action, or a block falling after being released, follows from the action history via learned physics.

  \item \textbf{Relative gripper-to-target distance:} The decreasing distance between the gripper and the target object is explained by the trajectory of past actions and the language goal.
\end{enumerate}

\textbf{Predictable but action-irrelevant content} (a subtlety). Not all predictable information is action-relevant. Static background texture (e.g., wood grain on the table) or a clock ticking on the wall are perfectly predictable from history---the prior loss does not penalize them. However, finite latent capacity creates an indirect pressure: channels spent encoding action-irrelevant background are channels unavailable for gripper geometry. The action latent loss $\loss_b$ amplifies this pressure, because the latent transition $\vz_t \to \vz_{t+1}$ must be informative enough to denoise the action $\vb_{t,\text{clean}}$, actively rewarding action-relevant structure over inert background content.

\medskip

The \emph{joint} denoising of $(\vb_{t,\text{clean}}, \vz_{t+1,\text{clean}})$ creates two complementary pressures that together favor action-relevant content:
\begin{itemize}[leftmargin=*,itemsep=2pt]
  \item \textbf{Forward dynamics:} The prior must denoise $\vz_{t+1}$---what will happen after executing the action. This requires the observation latent to encode physical geometry that actions causally affect (e.g., gripper-object contact, not table texture).
  \item \textbf{Inverse dynamics:} The prior must also denoise $\vb_{t,\text{clean}}$---what action caused the transition $\vz_t \to \vz_{t+1}$. This prevents the observation encoder from collapsing to a trivially predictable but uninformative representation (e.g., encoding only static background). The action encoder $A_\theta$ must simultaneously encode actions in a form that the prior can consistently denoise from the latent history.
\end{itemize}

\textbf{Formal argument.}
The total information in the latent can be decomposed as:
\begin{equation}
  I(\vz_{t+1,\text{clean}}; \vo_{t+1}) = \underbrace{I(\vz_{t+1,\text{clean}}; \vo_{t+1} \mid \mathcal{H}_t^z)}_{\text{unpredictable info (penalized by } \loss_z^{\text{latent}}\text{)}} + \underbrace{I(\vz_{t+1,\text{clean}}; \mathcal{H}_t^z)}_{\text{history-predictable info (free)}}.
  \label{eq:info_decomp}
\end{equation}
The flow matching prior loss $\loss_z^{\text{latent}}$ is a reweighted approximation of the KL-optimal bound on the first term (see \Cref{sec:theory} for proof that the information bottleneck property is preserved). The decoder loss $\loss_x$ incentivizes maximizing total information. The action latent loss $\loss_b$ ensures the latent captures the causal effect of actions (inverse dynamics signal) via the same flow matching objective, and creates capacity pressure that favors action-relevant content among the history-predictable remainder.

Together, three pressures shape the latent space: (1)~the prior loss \emph{discards} unpredictable content, (2)~the decoder loss \emph{retains} content useful for reconstruction, and (3)~the action loss \emph{prioritizes} action-relevant structure by requiring the latent transition to support action denoising.

\subsection{Deterministic Decoder for Future Frame Reconstruction}
\label{sec:decoder}

Unlike the original UL framework which uses a diffusion decoder, we use the \emph{pretrained deterministic Wan2.1 VAE decoder} $D_\theta$ to reconstruct observations from latents.
This is a practical choice: the Wan2.1 VAE already provides a high-quality deterministic decoder, and training a diffusion decoder from scratch would negate the benefits of pretraining.

Given the clean future latent $\vz_{t+1,\text{clean}} = E_\theta(\vo_{t+1})$, the decoder reconstructs:
\begin{equation}
  \hat{\vo}_{t+1} = D_\theta(\vz_{t+1,\text{clean}}).
  \label{eq:deterministic_decode}
\end{equation}

The reconstruction loss combines pixel-level and perceptual terms:
\begin{equation}
  \loss_x(\theta) = \|\vo_{t+1} - \hat{\vo}_{t+1}\|^2 + \lambda_{\text{perc}} \cdot \loss_{\text{perc}}(\vo_{t+1}, \hat{\vo}_{t+1}),
  \label{eq:recon_loss}
\end{equation}
where $\loss_{\text{perc}}$ is a perceptual loss (e.g., LPIPS) and $\lambda_{\text{perc}}$ controls its weight.

\textbf{Relation to the UL bound.}
In the original UL formulation, the diffusion decoder provides a tight variational bound on $-\log p_\theta(\vo_{t+1} \mid \vz_{t+1,0})$.
With a deterministic decoder, this bound becomes approximate: $\loss_x$ serves as a proxy for the decoder term in the ELBO, ensuring the latent retains sufficient information for reconstruction without the formal diffusion bound.
The loss factor $c_{\text{lf}} \in [1.3, 1.7]$ controls the reconstruction-compression trade-off, analogous to its role in UL.

\subsection{Action Decoder}
\label{sec:action_decoder}

To recover the raw action $\va_t \in \R^{d_a}$ from the denoised action latent $\hat{\vb}_t$, we use a lightweight action decoder:
\begin{equation}
  \hat{\va}_t = D_\theta^a(\hat{\vb}_t) \in \R^{d_a}.
  \label{eq:action_decode}
\end{equation}
The action decoder is a 2-layer MLP trained with an auxiliary reconstruction loss:
\begin{equation}
  \loss_a^{\text{recon}}(\theta) = \|\va_t - D_\theta^a(\vb_{t,\text{clean}})\|^2.
  \label{eq:action_recon_loss}
\end{equation}
This loss is applied to the clean action latent $\vb_{t,\text{clean}}$ and ensures that the action encoder-decoder pair faithfully represents the action space.

\subsection{Complete Stage 1 Objective}
\label{sec:stage1_objective}

The total per-transition loss combines the joint prior loss, the decoder loss, and the action reconstruction loss:
\begin{equation}
  \loss_{\text{S1}}(\theta) = \underbrace{\loss_z^{\text{latent}}(\theta) + \gamma \cdot \loss_b(\theta)}_{\text{joint dynamics prior (flow matching)}} + c_{\text{lf}} \cdot \underbrace{\loss_x(\theta)}_{\text{decoder}} + \underbrace{\loss_a^{\text{recon}}(\theta)}_{\text{action reconstruction}}.
  \label{eq:total_loss}
\end{equation}

\textbf{Interpretation of each term:}
\begin{itemize}[leftmargin=*,itemsep=2pt]
  \item $\loss_z^{\text{latent}}$: Flow matching velocity prediction loss for the observation latent. A reweighted approximation of the DDPM KL bound that penalizes unpredictable information in the future latent given the history. Forces the observation encoder to discard action-irrelevant content.
  \item $\loss_b$: Flow matching velocity prediction loss for the action latent. Uses the same objective as $\loss_z^{\text{latent}}$. Requires the latent history to contain enough dynamics structure that the action can be denoised. Prevents trivial latent collapse.
  \item $\loss_x$: Observation reconstruction loss (MSE + perceptual). Rewards information that aids reconstruction via the pretrained deterministic Wan decoder. Ensures the latent retains fine geometric detail (e.g., gripper-object distance).
  \item $\loss_a^{\text{recon}}$: Action reconstruction loss. Ensures the action encoder-decoder pair faithfully represents actions.
\end{itemize}

\subsection{Stage 1 Training Algorithm}
\label{sec:stage1_algo}

The complete Stage~1 training procedure is given in \Cref{alg:stage1}.

\begin{algorithm}[t]
\caption{Stage 1: Training AC-UL with Joint Dynamics Prior}
\label{alg:stage1}
\begin{algorithmic}[1]
\REQUIRE Dataset of trajectories $\{(\bm{l}, \vo_0, \va_0, \vo_1, \va_1, \ldots)\}$, history length $H$
\STATE Sample window: $(\bm{l}, \vo_{t-H+1:t+1}, \va_{t-H:t}) \sim \mathcal{D}$
\STATE \textbf{// Encode observations and actions}
\STATE Encode observations: $\vz_{k,\text{clean}} = E_\theta(\vo_k)$ for $k = t{-}H{+}1, \ldots, t{+}1$
\STATE Encode actions: $\vb_{k,\text{clean}} = A_\theta(\va_k)$ for $k = t{-}H, \ldots, t$
\STATE Construct latent history: $\mathcal{H}_t^z = (\bm{l}, \vz_{t-H+1,\text{clean}}, \vb_{t-H,\text{clean}}, \ldots, \vb_{t-1,\text{clean}}, \vz_{t,\text{clean}})$
\STATE \textbf{// Joint dynamics prior loss (flow matching)}
\STATE Sample $\tau \sim p(\tau)$, $\vep_z \sim \mathcal{N}(\bm{0}, \bm{I})$, $\vep_b \sim \mathcal{N}(\bm{0}, \bm{I})$
\STATE $\vz_{t+1,\tau} = (1-\tau)\,\vz_{t+1,\text{clean}} + \tau\,\vep_z$ \hfill \COMMENT{flow matching interpolation}
\STATE $\vb_{t,\tau} = (1-\tau)\,\vb_{t,\text{clean}} + \tau\,\vep_b$
\STATE $(v_\theta^b, v_\theta^z) = \hat{f}_\theta^{\text{dyn}}(\vb_{t,\tau}, \vz_{t+1,\tau}, \mathcal{H}_t^z, \tau)$ \hfill \COMMENT{velocity prediction}
\STATE $\loss_z^{\text{latent}} = \|v_\theta^z - (\vz_{t+1,\text{clean}} - \vep_z)\|^2$
\STATE $\loss_b = \|v_\theta^b - (\vb_{t,\text{clean}} - \vep_b)\|^2$
\STATE \textbf{// Action reconstruction loss}
\STATE $\loss_a^{\text{recon}} = \|\va_t - D_\theta^a(\vb_{t,\text{clean}})\|^2$
\STATE \textbf{// Reconstruction loss (deterministic decoder)}
\STATE $\hat{\vo}_{t+1} = D_\theta(\vz_{t+1,\text{clean}})$
\STATE $\loss_x = \|\vo_{t+1} - \hat{\vo}_{t+1}\|^2 + \lambda_{\text{perc}} \cdot \loss_{\text{perc}}(\vo_{t+1}, \hat{\vo}_{t+1})$
\STATE \textbf{// Combined loss}
\STATE $\loss_{\text{S1}} = \loss_z^{\text{latent}} + \gamma \cdot \loss_b + c_{\text{lf}} \cdot \loss_x + \loss_a^{\text{recon}}$
\STATE Update $\theta$ by gradient descent on $\loss_{\text{S1}}$
\end{algorithmic}
\end{algorithm}

\subsection{Architecture Details}
\label{sec:architecture}

\textbf{Observation encoder $E_\theta$ and decoder $D_\theta$.}
We initialize $E_\theta$ and $D_\theta$ from the pretrained Wan2.1 VAE~\citep{wan2025}, which encodes $320 \times 176$ RGB frames into $c = 16$ channel latents at $8\times$ spatial downsampling.
The decoder is deterministic (not a diffusion decoder): $\hat{\vo} = D_\theta(\vz_{\text{clean}})$.
Multi-view observations (e.g., three camera views in DROID) are composited into a single image following DreamZero's convention.
During Stage~1 fine-tuning, the VAE encoder and decoder parameters are updated via LoRA~\citep{hu2022lora} adapters (rank $r = 4$) to adapt the latent space while preserving pretrained visual features.

\textbf{Action encoder $A_\theta$.}
An embodiment-specific MLP (following DreamZero's \texttt{MultiEmbodimentActionEncoder} design) mapping raw actions $\va_k \in \R^{d_a}$ to action latents $\vb_{k,\text{clean}} \in \R^{d_b}$ with $d_b = 64$.
Actions are first normalized to $[-1, 1]$ using per-dataset quantile statistics.

\textbf{Action decoder $D_\theta^a$.}
An embodiment-specific MLP that maps denoised action latents $\hat{\vb}_t \in \R^{d_b}$ back to raw actions $\hat{\va}_t \in \R^{d_a}$.

\textbf{State encoder.}
An embodiment-specific MLP that encodes proprioceptive state $\bm{s}_t \in \R^{d_s}$ into a single conditioning token of dimension $d_b$.

\textbf{Joint dynamics prior $\hat{f}_\theta^{\text{dyn}}$.}
A causal diffusion transformer (DiT), initialized from a pretrained Wan2.1 video generation model at 1.3B parameter scale and fine-tuned with LoRA.
Initializing from pretrained Wan weights provides a strong starting point for dynamics modeling, as the pretrained model already captures visual transition priors from large-scale video data.
The input sequence follows DreamZero's blockwise structure: language tokens, followed by blocks of interleaved observation and action latent tokens, plus state tokens:
\begin{equation}
  \big[\text{text\_enc}(\bm{l}),\; \text{patchify}(\vz_{t-H+1}),\; \text{mlp}(\vb_{t-H}),\; \text{mlp}(\bm{s}),\; \ldots,\; \text{patchify}(\vz_{t}),\; \text{mlp}(\vb_{t,\tau}),\; \text{patchify}(\vz_{t+1,\tau})\big].
  \label{eq:prior_input}
\end{equation}
The language instruction $\bm{l}$ is encoded by a frozen text encoder (UMT5-XXL). Each observation latent $\vz_{k,\text{clean}}$ is patchified into spatial tokens; each action latent $\vb_{k,\text{clean}}$ and state token are projected through MLPs. The noised targets $\vb_{t,\tau}$ and $\vz_{t+1,\tau}$ are appended. The flow matching timestep $\tau$ is encoded via sinusoidal embeddings and added to the noisy target tokens.

The model uses \emph{blockwise causal attention} (following DreamZero): within each block, action tokens can attend to the current and all previous observation tokens, the current state token, and themselves. The model has two output heads that predict velocity fields:
\begin{itemize}[leftmargin=*,itemsep=2pt]
  \item \textbf{Observation latent head:} Linear projection from the output tokens at the $\vz_{t+1,\tau}$ positions $\to v_\theta^z \in \R^{h \times w \times c}$. The clean prediction is recovered as $\hat{\vz}_{t+1,\text{clean}} = \vz_{t+1,\tau} + \tau\, v_\theta^z$.
  \item \textbf{Action latent head:} Linear projection from the output tokens at the $\vb_{t,\tau}$ position $\to v_\theta^b \in \R^{d_b}$. The clean prediction is recovered as $\hat{\vb}_t = \vb_{t,\tau} + \tau\, v_\theta^b$.
\end{itemize}

%=====================================================================
\section{Stage 2: Autoregressive Diffusion Transformer on Frozen Latents}
\label{sec:stage2}
%=====================================================================

Once Stage~1 converges, we freeze the fine-tuned observation encoder $E_\theta$, action encoder $A_\theta$, observation decoder $D_\theta$, and action decoder $D_\theta^a$. The frozen encoders now provide a \emph{physics-grounded vocabulary} whose latent space has been adapted to contain action-predictable geometry.

For Stage~2, we train the full-scale autoregressive diffusion transformer (CausalWanModel, 1.3B parameters) on the frozen AC-UL latent space.
The Stage~2 DiT is also initialized from pretrained Wan2.1 weights and fine-tuned with LoRA (rank $r = 4$), following the DreamZero training recipe but operating on action-optimized latents instead of generic VAE latents.
This pretrained initialization is critical: training a 1.3B DiT from scratch on DROID alone (${\sim}76$K trajectories) would be insufficient, whereas LoRA fine-tuning adapts the pretrained video generation prior to the AC-UL latent space with modest data requirements.

\subsection{Latent Trajectory Extraction}

Given a trajectory $(\vo_0, \va_0, \vo_1, \va_1, \ldots, \vo_T)$, we extract clean latents for both observations and actions:
\begin{align}
  \vz_{k,\text{clean}} &= E_\theta(\vo_k), \quad k = 0, 1, \ldots, T, \label{eq:latent_extract} \\
  \vb_{k,\text{clean}} &= A_\theta(\va_k), \quad k = 0, 1, \ldots, T-1. \label{eq:action_latent_extract}
\end{align}

The trajectory in latent space becomes:
\begin{equation}
  (\vz_{0,\text{clean}}, \vb_{0,\text{clean}}, \vz_{1,\text{clean}}, \vb_{1,\text{clean}}, \ldots, \vz_{T,\text{clean}}).
  \label{eq:latent_trajectory}
\end{equation}

\subsection{Joint Video-Action Generation in Latent Space}

The Stage~2 model $\mathcal{T}_\phi$ is a diffusion transformer that takes the latent history and jointly generates the future latent and current action. We follow a chunk-wise autoregressive scheme inspired by DreamZero.

\textbf{Chunk definition.}
The trajectory is divided into chunks of $K$ latent frames. Each chunk $k$ contains interleaved observation and action latents:
\begin{equation}
  \mathcal{C}_k = \big(\vz_{kK+1,\text{clean}}, \vb_{kK,\text{clean}}, \vz_{kK+2,\text{clean}}, \vb_{kK+1,\text{clean}}, \ldots, \vz_{(k+1)K,\text{clean}}, \vb_{(k+1)K-1,\text{clean}}\big).
  \label{eq:chunk}
\end{equation}

\textbf{Flow-matching training.}
For each chunk, we sample noise and construct interpolants for both the observation latents and action latents:
\begin{align}
  \vz_{kK+j,\tau_k} &= (1-\tau_k)\,\vz_{kK+j,\text{clean}} + \tau_k\,\vep_z^{(j)}, \label{eq:flow_latent} \\
  \vb_{kK+j-1,\tau_k} &= (1-\tau_k)\,\vb_{kK+j-1,\text{clean}} + \tau_k\,\vep_b^{(j)}, \label{eq:flow_action}
\end{align}
where $\tau_k \sim \mathcal{U}(0,1)$ is sampled independently per chunk. The model predicts the velocity field:
\begin{equation}
  \loss_{\text{S2}} = \E\!\left[\frac{1}{K}\sum_{k} w(\tau_k)\,\left\|
    \mathcal{T}_\phi\!\left([\vz_{\tau_k}^k, \vb_{\tau_k}^k]; \mathcal{C}_{<k}, \bm{l}, \bm{s}, \tau_k\right)
    - [\vz_{\text{clean}}^k - \vep_z^k,\; \vb_{\text{clean}}^k - \vep_b^k]
  \right\|^2\right],
  \label{eq:stage2_flow}
\end{equation}
where $\mathcal{C}_{<k}$ denotes the clean context from previous chunks, $\bm{l}$ is the language instruction embedding, and $\bm{s}$ is the proprioceptive state. Note that both observation and action latents are denoised with the same flow-matching objective.

\textbf{Teacher forcing.}
During training, the model denoises the current chunk conditioned on \emph{clean} previous chunks via attention masking. This mirrors DreamZero's training but operates on compact AC-UL latents rather than high-dimensional video tokens and raw actions.

\subsection{Closed-Loop Inference}
\label{sec:inference}

During deployment, the model operates in a closed loop. At each step $t$, the history $\mathcal{H}_t^z$ contains the language instruction, past observation latents $\vz_{t-H+1:t,\text{clean}}$, and past action latents $\vb_{t-H:t-1,\text{clean}}$:

\begin{enumerate}[leftmargin=*,itemsep=2pt]
  \item \textbf{Encode:} $\vz_{t,\text{clean}} = E_\theta(\vo_t)$ --- encode the current observation with the frozen observation encoder.
  \item \textbf{Generate:} $(\hat{\vb}_t, \hat{\vz}_{t+1,\text{clean}}) = \mathcal{T}_\phi(\mathcal{H}_t^z)$ --- jointly denoise the action latent and observation latent via iterative denoising from noise.
  \item \textbf{Decode action:} $\hat{\va}_t = D_\theta^a(\hat{\vb}_t)$ --- decode the denoised action latent to a raw action using the frozen action decoder.
  \item \textbf{Execute:} Apply $\hat{\va}_t$ to the robot. This action \emph{causes} the transition to the next state.
  \item \textbf{Observe:} Receive the new observation $\vo_{t+1}$ (the actual result of $\hat{\va}_t$).
  \item \textbf{Replace:} Compute $\vz_{t+1,\text{clean}} = E_\theta(\vo_{t+1})$ and replace $\hat{\vz}_{t+1,\text{clean}}$ in the KV cache. Compute $\vb_{t,\text{clean}} = A_\theta(\hat{\va}_t)$ for the history. This prevents error accumulation.
  \item \textbf{Advance:} Shift the history window: $\vb_{t,\text{clean}}$ becomes part of the past action latents, $\vz_{t+1,\text{clean}}$ becomes the latest observation latent. Go to step 2 with $t \leftarrow t+1$.
\end{enumerate}

\subsection{Planning via Latent Rollouts}
\label{sec:planning}

For goal-conditioned tasks, the world model head enables planning without a separate planner. Given a goal observation $\vo_g$, we:

\begin{enumerate}[leftmargin=*,itemsep=2pt]
  \item Encode the goal: $\vz_{g,\text{clean}} = E_\theta(\vo_g)$.
  \item Use the joint model to autoregressively generate action-latent rollouts \emph{without} ground-truth replacement:
  \begin{equation}
    (\hat{\vb}_t, \hat{\vz}_{t+1}) \to (\hat{\vb}_{t+1}, \hat{\vz}_{t+2}) \to \cdots \to (\hat{\vb}_{t+P-1}, \hat{\vz}_{t+P}).
    \label{eq:open_loop_rollout}
  \end{equation}
  At each step, the predicted $\hat{\vz}$ and $\hat{\vb}$ are fed back as context for the next step.
  \item Score each rollout by the terminal distance: $d = \|\hat{\vz}_{t+P,\text{clean}} - \vz_{g,\text{clean}}\|^2$.
  \item Use CEM or best-of-$N$ sampling to select the best action sequence.
  \item Execute the first action and replan (receding horizon).
\end{enumerate}

Because the latents encode only action-relevant geometry, the planning cost function directly measures progress toward the goal in terms of physical state change.

%=====================================================================
\section{Theoretical Analysis}
\label{sec:theory}
%=====================================================================

\subsection{Flow Matching as an Approximation of the UL KL Bound}
\label{sec:fm_approx}

We now show formally that the flow matching prior loss used in AC-UL is a reweighted form of the original DDPM KL bound from UL, preserving the information bottleneck property.

\begin{proposition}[Velocity-to-Clean Equivalence]
\label{prop:fm_equiv}
For the flow matching interpolation $\vz_\tau = (1-\tau)\vz_{\text{clean}} + \tau\vep$ with velocity target $v_{\text{target}} = \vz_{\text{clean}} - \vep$, the velocity prediction MSE equals a $\tau$-reweighted clean-sample prediction MSE:
\begin{equation}
  \|v_\theta(\vz_\tau, \tau) - (\vz_{\text{clean}} - \vep)\|^2 = \frac{1}{\tau^2}\,\|\vz_{\text{clean}} - \hat{\vz}_\theta\|^2,
  \label{eq:fm_clean_equiv}
\end{equation}
where the implied clean-sample prediction is $\hat{\vz}_\theta \triangleq \vz_\tau + \tau\, v_\theta(\vz_\tau, \tau)$.
\end{proposition}

\begin{proof}
From $\vz_\tau = (1-\tau)\vz_{\text{clean}} + \tau\vep$, we have $\vz_{\text{clean}} = (\vz_\tau - \tau\vep)/(1-\tau)$. The velocity target is $v_{\text{target}} = (\vz_{\text{clean}} - \vz_\tau)/\tau + \vz_{\text{clean}} - \vz_{\text{clean}} = (\vz_{\text{clean}} - \vz_\tau)/\tau$.
More directly: $v_{\text{target}} = \vz_{\text{clean}} - \vep$, and $\hat{\vz}_\theta = \vz_\tau + \tau v_\theta$. Then:
\begin{align}
  \vz_{\text{clean}} - \hat{\vz}_\theta &= \vz_{\text{clean}} - \vz_\tau - \tau v_\theta = \vz_{\text{clean}} - (1-\tau)\vz_{\text{clean}} - \tau\vep - \tau v_\theta \nonumber \\
  &= \tau(\vz_{\text{clean}} - \vep) - \tau v_\theta = \tau(v_{\text{target}} - v_\theta).
\end{align}
Therefore $\|\vz_{\text{clean}} - \hat{\vz}_\theta\|^2 = \tau^2\|v_{\text{target}} - v_\theta\|^2$.
\end{proof}

\begin{proposition}[Flow Matching Approximates the KL Bound]
\label{prop:fm_kl}
The flow matching prior loss and the tight DDPM KL bound share the same denoising MSE kernel $\|\vz_{\text{clean}} - \hat{\vz}_\theta\|^2$, differing only in the weighting over $\tau$:
\begin{align}
  \loss_z^{\text{KL}} &= \E_\tau\!\left[w_{\KL}(\tau)\,\|\vz_{\text{clean}} - \hat{\vz}_\theta\|^2\right], \quad w_{\KL}(\tau) = \frac{1-\tau}{\tau^3}, \label{eq:ddpm_weight} \\
  \loss_z^{\text{FM}} &= \E_\tau\!\left[w_{\text{FM}}(\tau)\,\|\vz_{\text{clean}} - \hat{\vz}_\theta\|^2\right], \quad w_{\text{FM}}(\tau) = \frac{1}{\tau^2}. \label{eq:fm_weight}
\end{align}
The ratio $w_{\text{FM}}(\tau)/w_{\KL}(\tau) = \tau/(1-\tau)$ is bounded, so the flow matching loss is a consistent proxy for the KL bound.
\end{proposition}

\begin{proof}
For the flow matching interpolation $\alpha(\tau) = 1-\tau$, $\sigma(\tau) = \tau$, the log-SNR is $\lambda(\tau) = 2\log\!\big(\frac{1-\tau}{\tau}\big)$. Computing:
$\frac{d\lambda}{d\tau} = \frac{-2}{\tau(1-\tau)}, \quad \exp\lambda(\tau) = \frac{(1-\tau)^2}{\tau^2}.$
The KL-optimal weight is $w_{\KL}(\tau) = -\frac{d\lambda}{d\tau}\frac{e^\lambda}{2} = \frac{(1-\tau)}{\tau^3}$.
From \Cref{prop:fm_equiv}, the FM loss uses $w_{\text{FM}}(\tau) = 1/\tau^2$.
\end{proof}

\textbf{Key implication.}
Since both losses are monotonically increasing in the clean-sample prediction error $\|\vz_{\text{clean}} - \hat{\vz}_\theta\|^2$ at every noise level $\tau$, they share the same \emph{information bottleneck property}: if the encoder $E_\theta$ encodes content into $\vz_{\text{clean}}$ that is unpredictable from the history $\mathcal{H}_t^z$, then $\|\vz_{\text{clean}} - \hat{\vz}_\theta\|^2$ increases (the prior cannot predict the unpredictable content), which increases the loss \emph{regardless of the weighting function}.
The flow matching loss differs from the KL bound only in how it distributes gradient signal across noise levels---it up-weights high-noise (large $\tau$) relative to the KL bound, which emphasizes low-noise prediction.

\subsection{AC-UL as an Approximate Variational Bound}

\begin{proposition}[AC-UL Approximate Bound]
The AC-UL Stage~1 loss provides an approximate upper bound on the negative log-likelihood of the joint action-observation transition:
\begin{equation}
  -\log p_\theta(\va_t, \vo_{t+1} \mid \mathcal{H}_t) \;\lessapprox\; \loss_z^{\text{latent}}(\theta) + \gamma \cdot \loss_b(\theta) + c_{\text{lf}} \cdot \loss_x(\theta) + \loss_a^{\text{recon}}(\theta),
  \label{eq:acul_bound}
\end{equation}
where $\lessapprox$ denotes that the bound is approximate due to three relaxations: (1)~the flow matching prior loss uses $w_{\text{FM}}$ instead of $w_{\KL}$, (2)~the deterministic decoder loss $\loss_x$ replaces the diffusion decoder bound, and (3)~the deterministic encoder lacks the minimum noise required for a finite KL, so the formal ELBO is not directly available (see below).
\end{proposition}

\textbf{Connection to the ELBO.}
If one were to use a DDPM noise schedule with minimum noise $\sigma(0) > 0$ (as in the original UL), the ELBO decomposition would be exact:
\begin{align}
  -\log p_\theta(\va_t, \vo_{t+1} \mid \mathcal{H}_t)
  &\leq \underbrace{-\E\!\Big[\log p_\theta(\vo_{t+1} \mid \vz_{t+1,0})\Big]}_{\text{decoder term}} + \underbrace{\KL\!\Big[q(\vz_{t+1,0}) \;\Big\|\; p_\theta(\vz_{t+1,0} \mid \mathcal{H}_t^z)\Big]}_{\text{obs.\ prior KL}} \nonumber \\
  &\quad + \underbrace{\KL\!\Big[q(\vb_{t,0}) \;\Big\|\; p_\theta(\vb_{t,0} \mid \mathcal{H}_t^z)\Big]}_{\text{action prior KL}} + \underbrace{-\E\!\Big[\log p_\theta(\va_t \mid \vb_{t,0})\Big]}_{\text{action decoder}},
  \label{eq:proof_step}
\end{align}
with each KL bounded by the DDPM denoising loss. In our flow matching formulation, the encoder is deterministic at $\tau = 0$ ($\vz_0 = \vz_{\text{clean}}$, a Dirac delta), so the KL terms are formally infinite and the ELBO does not hold as stated.
Instead, the FM loss serves as a \emph{practical proxy} for the KL terms: it shares the same denoising MSE kernel $\|\vz_{\text{clean}} - \hat{\vz}_\theta\|^2$ and increases monotonically whenever the encoder encodes unpredictable content, preserving the information bottleneck mechanism (\Cref{prop:fm_kl}).
The decoder and action decoder terms are bounded by $\loss_x$ and $\loss_a^{\text{recon}}$ respectively.

\subsubsection{Summary of Approximations}

Both the ideal UL objective and the AC-UL objective are weighted integrals of the same per-$\tau$ prediction error $L(\tau) \triangleq \|\vz_{\text{clean}} - \hat{\vz}_\theta(\vz_\tau, \mathcal{H}_t^z, \tau)\|^2$.
We summarize the three gaps between them:

\begin{center}
\begin{tabular}{p{2.2cm}p{3.3cm}p{3.3cm}p{4.5cm}}
\toprule
\textbf{Component} & \textbf{Ideal UL} & \textbf{AC-UL} & \textbf{Effect} \\
\midrule
Prior weighting & $w_{\KL}(\tau) = \frac{1-\tau}{\tau^3}$ (tight KL bound) & $w_{\text{FM}}(\tau) = \frac{1}{\tau^2}$ (flow matching) & Different gradient distribution across $\tau$; same minimizer $\theta^*$ \\[4pt]
Decoder & Diffusion decoder (tight $-\log p_\theta(\vo|\vz_0)$ bound) & Deterministic MSE + perceptual loss & Approximate reconstruction term \\[4pt]
Encoder & Stochastic ($\sigma(0){>}0$, finite KL) & Deterministic ($\vz_0 {=} \vz_{\text{clean}}$, KL $= \infty$) & No formal ELBO; FM loss is a practical proxy \\
\bottomrule
\end{tabular}
\end{center}

\textbf{Why the bottleneck is preserved despite these gaps.}
The key observation is that $L(\tau) \geq 0$ and both $w_{\KL}(\tau), w_{\text{FM}}(\tau) > 0$ for all $\tau \in (0,1)$.
Three properties follow:
\begin{enumerate}[leftmargin=*,itemsep=2pt]
  \item \textbf{Same minimizer:} If $\hat{\vz}_\theta$ equals the optimal predictor $\E[\vz_{\text{clean}} \mid \vz_\tau, \mathcal{H}_t^z]$ at every $\tau$, then $L(\tau)$ is minimized simultaneously at all noise levels. Both $\loss^{\KL}$ and $\loss^{\text{FM}}$ reach their minimum at the same $\theta^*$.
  \item \textbf{Same gradient direction:} At each $\tau$, both losses compute $\nabla_\theta L(\tau)$ scaled by a positive constant. Both push the encoder in the same direction.
  \item \textbf{Monotonic penalty:} If the encoder encodes unpredictable content, $L(\tau)$ increases at some $\tau$. Since both weightings are positive, both losses increase. The information bottleneck works with \emph{any} positive weighting function.
\end{enumerate}

The weighting ratio $w_{\text{FM}}(\tau)/w_{\KL}(\tau) = \tau/(1-\tau)$ shows that FM underweights low-noise prediction ($\tau \to 0$, ratio $\to 0$) and overweights high-noise denoising ($\tau \to 1$, ratio $\to \infty$), with equality at $\tau = 0.5$.
For the information bottleneck, this difference is immaterial: the bottleneck only requires that encoding unpredictable content increases $L(\tau)$ at some $\tau > 0$, which any positive weighting captures.

\subsection{Information-Theoretic Interpretation}

\textbf{Rate (prior loss).}
The prior loss $\loss_z^{\text{latent}}$ provides a reweighted bound on the mutual information between the encoder output and the data, \emph{beyond what is predictable from the history}:
\begin{equation}
  \loss_z^{\text{latent}} \propto I(\vz_{t+1,\text{clean}}; \vo_{t+1} \mid \mathcal{H}_t^z),
  \label{eq:rate_bound}
\end{equation}
where the proportionality reflects the reweighting from $w_{\KL}$ to $w_{\text{FM}}$. The \emph{direction} of the incentive is preserved: the loss increases monotonically with unpredictable information, regardless of the weighting.

\textbf{Action regularization.}
The action latent loss $\loss_b$ provides an additional inductive bias via the same flow matching objective. The prior must predict the action velocity from the latent history, so the latent space must encode the \emph{causal effect} of the action on the scene:
\begin{equation}
  \loss_b \propto I(\vb_{t,\text{clean}}; \va_t \mid \mathcal{H}_t^z),
  \label{eq:action_info}
\end{equation}
ensuring the latent transition $\vz_t \to \vz_{t+1}$ remains informative about the action.

\textbf{Equilibrium.}
At convergence, both encoders achieve an optimal trade-off: they maximize history-predictable, action-relevant information while minimizing unpredictable content:
\begin{equation}
  \vz_{t+1,\text{clean}}^* = \arg\min_{\vz} \Big[\underbrace{I(\vz; \vo_{t+1} \mid \mathcal{H}_t^z)}_{\text{obs.\ rate}} + c_{\text{lf}} \cdot \underbrace{D(\vo_{t+1}, \hat{\vo}_\theta)}_{\text{distortion}} - \gamma \cdot \underbrace{I(\vb_{t,\text{clean}}; \vz \mid \mathcal{H}_t^z)}_{\text{action-obs coupling}}\Big].
  \label{eq:rd_tradeoff}
\end{equation}

%=====================================================================
\section{Experiments}
\label{sec:experiments}
%=====================================================================

We evaluate AC-UL across three experimental settings: (1)~closed-loop planning via latent rollouts, (2)~imitation learning with joint world-action modeling, and (3)~real-world video prediction.

\subsection{Datasets and Benchmarks}
\label{sec:datasets}

\textbf{DROID}~\citep{khazatsky2024droid}.
Our primary dataset. A large-scale real-world manipulation dataset with ${\sim}76$K trajectories across 564 scenes on Franka Panda robots. Provides multi-view RGB from 3 cameras (wrist, left exterior, right exterior) at $320 \times 240$, 7-DoF joint position actions plus gripper ($d_a = 8$), proprioceptive state ($d_s = 8$), and language annotations.
Multi-view images are composited into a single $320 \times 176$ image following DreamZero's convention.
We use DROID for all main experiments: Stage~1 fine-tuning, Stage~2 training, and evaluation.

\textbf{Bridge-v2}~\citep{walke2023bridgedata}.
A real-world dataset with ${\sim}60$K trajectories on a WidowX robot with 7-DoF actions across kitchen tasks. Used for cross-embodiment transfer experiments.

\subsection{Baselines}
\label{sec:baselines}

\begin{itemize}[leftmargin=*,itemsep=2pt]
  \item \textbf{DreamZero (14B)}~\citep{dreamzero2025}: The full-scale DreamZero model. Autoregressive flow-matching transformer (14B parameters, LoRA fine-tuned) for joint video-action generation on frozen Wan2.1 VAE latents. This is the primary baseline: we aim to match or exceed its performance with a $10\times$ smaller model.
  \item \textbf{DreamZero (1.3B, frozen VAE)}: DreamZero at the same 1.3B scale as AC-UL, using frozen Wan2.1 VAE latents without AC-UL fine-tuning. This controls for model scale and isolates the effect of latent space adaptation.
  \item \textbf{DINO-WM}~\citep{zhou2024dinowm}: Frozen DINOv2 features with ViT transition model and CEM planning.
  \item \textbf{FLARE}~\citep{nvidia2025flare}: Diffusion transformer VLA with future latent alignment using SigLIP-2 features.
  \item \textbf{TD-MPC2}~\citep{hansen2024tdmpc2}: Decoder-free latent dynamics model with temporal difference learning.
\end{itemize}

The core experimental narrative is a three-way comparison:
\begin{center}
\begin{tabular}{lcc}
\toprule
\textbf{Model} & \textbf{Params} & \textbf{Latent Space} \\
\midrule
DreamZero & 14B & Frozen Wan VAE \\
DreamZero (1.3B) & 1.3B & Frozen Wan VAE \\
AC-UL (ours) & 1.3B & Fine-tuned AC-UL \\
\bottomrule
\end{tabular}
\end{center}
Comparing the last two rows isolates the latent space effect at equal scale; comparing the first and last rows tests whether better representations can compensate for $10\times$ fewer parameters.

\subsection{Experiment 1: Closed-Loop Planning via Latent Rollouts}
\label{sec:exp_planning}

\subsubsection{Setup}

We evaluate goal-conditioned planning using the Stage~2 model's latent rollout capability (\Cref{sec:planning}). The agent receives a current observation $\vo_t$ and a goal image $\vo_g$, and must generate action sequences that reach the goal.

\textbf{Planning method.}
We use best-of-$N$ sampling: generate $N = 256$ latent-action rollouts from the joint model, score each by terminal distance to the goal latent, select the best trajectory, execute the first $K$ actions, and replan.

\textbf{Metrics.}
Success rate (SR) for discrete tasks; LPIPS between achieved and goal observations.

\subsubsection{Results}

\begin{table}[t]
\centering
\caption{Goal-conditioned planning performance on DROID (Success Rate \%). AC-UL (1.3B) generates action sequences via latent rollouts from the joint video-action model.}
\label{tab:planning}
\begin{tabular}{llccccc}
\toprule
\textbf{Method} & \textbf{Params} & \textbf{Pick} & \textbf{Place} & \textbf{Stack} & \textbf{Pour} & \textbf{Avg.} \\
\midrule
DreamZero (14B, frozen VAE)   & 14B  & -- & -- & -- & -- & -- \\
DreamZero (1.3B, frozen VAE)  & 1.3B & -- & -- & -- & -- & -- \\
DINO-WM                       & --   & -- & -- & -- & -- & -- \\
TD-MPC2                       & --   & -- & -- & -- & -- & -- \\
\midrule
AC-UL (ours)                  & 1.3B & -- & -- & -- & -- & -- \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[t]
\centering
\caption{Planning with visual generalization on DROID: novel scenes and object configurations not seen during training. Success Rate (\%).}
\label{tab:planning_generalization}
\begin{tabular}{llccc}
\toprule
\textbf{Method} & \textbf{Params} & \textbf{Novel Scene} & \textbf{Novel Objects} & \textbf{Avg.} \\
\midrule
DreamZero (14B, frozen VAE)   & 14B  & -- & -- & -- \\
DreamZero (1.3B, frozen VAE)  & 1.3B & -- & -- & -- \\
DINO-WM                       & --   & -- & -- & -- \\
\midrule
AC-UL (ours)                  & 1.3B & -- & -- & -- \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Experiment 2: Imitation Learning with Joint World-Action Modeling}
\label{sec:exp_imitation}

\subsubsection{Setup}

We compare AC-UL's Stage~2 policy (the action output from the joint model) against FLARE and other VLA baselines on imitation learning benchmarks.

\textbf{Architecture.}
Stage~2 uses the Wan2.1 1.3B CausalWanModel (initialized from pretrained weights, LoRA fine-tuned), chunk size $K = 2$ latent frames. Language instructions are encoded with a frozen text encoder (UMT5-XXL). Flow-matching with $K_{\text{denoise}} = 4$ denoising steps at inference.

\textbf{Closed-loop evaluation.}
The model generates actions in closed loop: at each step, the predicted latent is replaced with the ground-truth encoding of the new observation in the KV cache.

\textbf{Metrics.}
Task success rate averaged over 3 seeds with 20 evaluation episodes per task.

\subsubsection{Results}

\begin{table}[t]
\centering
\caption{Imitation learning performance on DROID (Success Rate \%). AC-UL Stage~2 generates actions from the joint video-action model in closed loop.}
\label{tab:imitation}
\begin{tabular}{llccccc}
\toprule
\textbf{Method} & \textbf{Params} & \textbf{Pick} & \textbf{Place} & \textbf{Stack} & \textbf{Pour} & \textbf{Avg.} \\
\midrule
DreamZero (14B, frozen VAE)      & 14B  & -- & -- & -- & -- & -- \\
DreamZero (1.3B, frozen VAE)     & 1.3B & -- & -- & -- & -- & -- \\
FLARE                            & --   & -- & -- & -- & -- & -- \\
DINO-WM + BC                     & --   & -- & -- & -- & -- & -- \\
\midrule
AC-UL Stage~2 (ours)             & 1.3B & -- & -- & -- & -- & -- \\
AC-UL + planning (ours)          & 1.3B & -- & -- & -- & -- & -- \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[t]
\centering
\caption{Long-horizon evaluation on DROID. Success Rate (\%) for multi-step task chains. Language-conditioned.}
\label{tab:long_horizon}
\begin{tabular}{llccccc}
\toprule
\textbf{Method} & \textbf{Params} & \textbf{1 step} & \textbf{2 steps} & \textbf{3 steps} & \textbf{4 steps} & \textbf{5 steps} \\
\midrule
DreamZero (14B, frozen VAE)   & 14B  & -- & -- & -- & -- & -- \\
DreamZero (1.3B, frozen VAE)  & 1.3B & -- & -- & -- & -- & -- \\
FLARE                         & --   & -- & -- & -- & -- & -- \\
\midrule
AC-UL Stage~2 (ours)          & 1.3B & -- & -- & -- & -- & -- \\
\bottomrule
\end{tabular}
\end{table}

\subsubsection{Data Efficiency}

\begin{table}[t]
\centering
\caption{Data efficiency on DROID. Success Rate (\%) with varying amounts of training data (\% of full DROID dataset).}
\label{tab:data_efficiency}
\begin{tabular}{llcccc}
\toprule
\textbf{Method} & \textbf{Params} & \textbf{10\%} & \textbf{25\%} & \textbf{50\%} & \textbf{100\%} \\
\midrule
DreamZero (14B, frozen VAE)   & 14B  & -- & -- & -- & -- \\
DreamZero (1.3B, frozen VAE)  & 1.3B & -- & -- & -- & -- \\
FLARE                         & --   & -- & -- & -- & -- \\
\midrule
AC-UL Stage~2 (ours)          & 1.3B & -- & -- & -- & -- \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Experiment 3: Real-World Video-Action Prediction}
\label{sec:exp_video}

\subsubsection{Setup}

We evaluate the quality of joint video-action prediction on the DROID dataset. We train Stage~1 on DROID training split, then evaluate:
\begin{itemize}[leftmargin=*,itemsep=2pt]
  \item \textbf{Latent prediction accuracy:} MSE between predicted and ground-truth latents over multi-step rollouts (1, 5, 10, 20 steps).
  \item \textbf{Action prediction accuracy:} MSE between predicted and ground-truth actions.
  \item \textbf{Decoded video quality:} FVD, LPIPS, SSIM, PSNR of decoded predictions vs.\ ground truth.
  \item \textbf{Action-relevance metric:} Correlation between $\|\Delta \vz_k\|$ and $\|\va_k\|$.
\end{itemize}

\subsubsection{Results}

\begin{table}[t]
\centering
\caption{Video prediction quality on DROID test set. Multi-step prediction from ground-truth initial observation and open-loop rollout.}
\label{tab:video_pred}
\begin{tabular}{llcccc}
\toprule
\textbf{Method} & \textbf{Params} & \textbf{FVD$\downarrow$} & \textbf{LPIPS$\downarrow$} & \textbf{SSIM$\uparrow$} & \textbf{PSNR$\uparrow$} \\
\midrule
DreamZero (14B, frozen VAE)   & 14B  & -- & -- & -- & -- \\
DreamZero (1.3B, frozen VAE)  & 1.3B & -- & -- & -- & -- \\
DINO-WM                       & --   & -- & -- & -- & -- \\
\midrule
AC-UL (ours)                  & 1.3B & -- & -- & -- & -- \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[t]
\centering
\caption{Latent prediction MSE and action prediction MSE over rollout horizon on DROID test set.}
\label{tab:latent_rollout}
\begin{tabular}{llcccc|cccc}
\toprule
& & \multicolumn{4}{c|}{\textbf{Latent MSE$\downarrow$}} & \multicolumn{4}{c}{\textbf{Action MSE$\downarrow$}} \\
\textbf{Method} & \textbf{Params} & \textbf{1} & \textbf{5} & \textbf{10} & \textbf{20} & \textbf{1} & \textbf{5} & \textbf{10} & \textbf{20} \\
\midrule
DreamZero (14B, frozen VAE)   & 14B  & -- & -- & -- & -- & -- & -- & -- & -- \\
DreamZero (1.3B, frozen VAE)  & 1.3B & -- & -- & -- & -- & -- & -- & -- & -- \\
DINO-WM                       & --   & -- & -- & -- & -- & -- & -- & -- & -- \\
\midrule
AC-UL (ours)                  & 1.3B & -- & -- & -- & -- & -- & -- & -- & -- \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Experiment 4: Cross-Embodiment Transfer}
\label{sec:exp_transfer}

\subsubsection{Setup}

We train Stage~1 on combined DROID (Franka) + Bridge-v2 (WidowX) data and evaluate whether a single encoder learns embodiment-agnostic physics representations.

\subsubsection{Results}

\begin{table}[t]
\centering
\caption{Cross-embodiment transfer. Stage~1 fine-tuned on combined DROID + Bridge-v2 data, Stage~2 trained on target embodiment. Success Rate (\%).}
\label{tab:cross_embodiment}
\begin{tabular}{llccc}
\toprule
\textbf{Method} & \textbf{Params} & \textbf{Franka (DROID)} & \textbf{WidowX (Bridge)} & \textbf{Avg.} \\
\midrule
DreamZero (14B, frozen VAE, shared)  & 14B  & -- & -- & -- \\
DreamZero (1.3B, frozen VAE, shared) & 1.3B & -- & -- & -- \\
DINO-WM (shared)                     & --   & -- & -- & -- \\
\midrule
AC-UL (shared, ours)                 & 1.3B & -- & -- & -- \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Ablation Studies}
\label{sec:ablations}

\subsubsection{Effect of History-Conditioned Joint Prior}

\begin{table}[t]
\centering
\caption{Ablation: prior conditioning variants on DROID. Planning SR (\%).}
\label{tab:ablation_prior}
\begin{tabular}{lcc}
\toprule
\textbf{Prior Variant} & \textbf{Planning SR} & \textbf{Bitrate (bpd)} \\
\midrule
Frozen VAE, no fine-tuning (DreamZero baseline)   & -- & -- \\
AC-UL, latent only: $\mathcal{H}_t^z \to \vz_{t+1}$ & -- & -- \\
AC-UL, joint (ours): $\mathcal{H}_t^z \to (\vz_{t+1}, \vb_{t,\text{clean}})$ & -- & -- \\
\bottomrule
\end{tabular}
\end{table}

\subsubsection{Effect of Action Loss Weight}

\begin{table}[t]
\centering
\caption{Ablation: action loss weight $\gamma$ on DROID.}
\label{tab:ablation_gamma}
\begin{tabular}{lccc}
\toprule
$\gamma$ & \textbf{Planning SR} & \textbf{Action MSE} & \textbf{Latent MSE} \\
\midrule
0.0 (no action loss)  & -- & -- & -- \\
0.1                    & -- & -- & -- \\
0.5                    & -- & -- & -- \\
1.0                    & -- & -- & -- \\
2.0                    & -- & -- & -- \\
\bottomrule
\end{tabular}
\end{table}

\subsubsection{Effect of History Length}

\begin{table}[t]
\centering
\caption{Ablation: history length $H$ on DROID. Planning SR (\%).}
\label{tab:ablation_history}
\begin{tabular}{lccc}
\toprule
$H$ & \textbf{Planning SR} & \textbf{Action MSE} & \textbf{Latent MSE} \\
\midrule
1 (Markov)  & -- & -- & -- \\
2           & -- & -- & -- \\
4           & -- & -- & -- \\
8           & -- & -- & -- \\
\bottomrule
\end{tabular}
\end{table}

\subsubsection{Effect of Loss Factor and Timestep Distribution}

\begin{table}[t]
\centering
\caption{Ablation: loss factor $c_{\text{lf}}$ and shifted Beta distribution shift parameter on DROID.}
\label{tab:ablation_lf_noise}
\begin{tabular}{lcccc}
\toprule
$c_{\text{lf}}$ & Shift & \textbf{Planning SR} & \textbf{PSNR} & \textbf{Bitrate} \\
\midrule
1.3 & 5.0   & -- & -- & -- \\
1.5 & 5.0   & -- & -- & -- \\
1.7 & 5.0   & -- & -- & -- \\
1.5 & 1.0   & -- & -- & -- \\
1.5 & 3.0   & -- & -- & -- \\
1.5 & 10.0  & -- & -- & -- \\
\bottomrule
\end{tabular}
\end{table}

\subsubsection{AC-UL Latents vs.\ Pixel Space for DreamZero}

\begin{table}[t]
\centering
\caption{Direct comparison: same 1.3B DiT architecture trained on different latent spaces. Same model size and training budget. DROID success rate (\%).}
\label{tab:ablation_latent_vs_pixel}
\begin{tabular}{lcccc}
\toprule
\textbf{Latent Space} & \textbf{Pick} & \textbf{Place} & \textbf{Avg.} & \textbf{Inference (ms)} \\
\midrule
Wan2.1 VAE (frozen)      & -- & -- & -- & -- \\
DINO features (frozen)   & -- & -- & -- & -- \\
AC-UL latents (ours)     & -- & -- & -- & -- \\
\bottomrule
\end{tabular}
\end{table}

%=====================================================================
\section{Analysis and Discussion}
\label{sec:analysis}
%=====================================================================

\subsection{What Do AC-UL Latents Encode?}

To verify that AC-UL latents focus on action-relevant geometry, we analyze:

\textbf{Latent sensitivity to actions.}
We compute the Jacobian $\partial \vz_{t+1,\text{clean}} / \partial \va_t$ and visualize which spatial positions are most sensitive to action perturbations. We expect high sensitivity near the gripper and manipulated objects.

\textbf{Latent change vs.\ action magnitude.}
We measure the Pearson correlation between $\|\Delta \vz_t\|$ and $\|\va_t\|$. High correlation confirms that latent dynamics are action-driven.

\textbf{Closed-loop vs.\ open-loop rollouts.}
Because AC-UL latents are compact and action-focused, we expect open-loop rollouts (without ground-truth replacement) to degrade more gracefully than pixel-space rollouts, as there is less irrelevant variation to compound.

\subsection{Comparison with DreamZero}

AC-UL and DreamZero share the same Wan-based architecture and the autoregressive joint video-action generation paradigm. The key differences are (1)~how the VAE latent space is obtained and (2)~model scale:
\begin{itemize}[leftmargin=*,itemsep=2pt]
  \item \textbf{DreamZero (14B)} uses a 14B CausalWanModel with the Wan2.1 VAE \emph{as-is}. The VAE was pretrained for generic video reconstruction, so its latent space retains textures, lighting, and background details irrelevant to action dynamics.
  \item \textbf{AC-UL (1.3B)} uses a $10\times$ smaller 1.3B CausalWanModel but \emph{fine-tunes} the VAE with the UL variational objective (implemented via flow matching), adding a dynamics prior that penalizes action-irrelevant information. The action encoder/decoder adds an explicit action latent channel with the same flow matching objective.
\end{itemize}
Our three-way comparison (14B frozen VAE, 1.3B frozen VAE, 1.3B AC-UL) isolates both effects: the 1.3B-vs-1.3B comparison shows the latent space improvement at equal scale, while the 1.3B AC-UL vs.\ 14B DreamZero comparison tests whether adapting the latent space can compensate for $10\times$ fewer parameters.

\subsection{Comparison with FLARE}

FLARE adds future latent prediction as an auxiliary loss using frozen SigLIP-2 features. AC-UL differs:
\begin{itemize}[leftmargin=*,itemsep=2pt]
  \item FLARE's targets (SigLIP-2) encode general visual semantics, not action-predictable physics.
  \item AC-UL's latents are \emph{trained} to contain only action-predictable geometry.
  \item AC-UL jointly generates observations and actions; FLARE generates only actions with future features as auxiliary.
\end{itemize}

\subsection{Limitations}

\begin{itemize}[leftmargin=*,itemsep=2pt]
  \item \textbf{Two-stage training:} Stage~1 requires fine-tuning the VAE and DiT (as dynamics prior) simultaneously with the action encoder/decoder, adding training cost over DreamZero's single-stage LoRA fine-tuning.
  \item \textbf{VAE fine-tuning risk:} Aggressive fine-tuning of the pretrained VAE could degrade general visual features. LoRA mitigates this but limits the degree of latent space adaptation.
  \item \textbf{Fixed encoders at Stage~2:} If Stage~1 fine-tuning is suboptimal, Stage~2 inherits these limitations.
  \item \textbf{No formal ELBO:} Three relaxations prevent a formal variational bound: (1)~the flow matching prior uses reweighted denoising ($w_{\text{FM}}$ instead of $w_{\KL}$), (2)~the deterministic Wan decoder replaces the diffusion decoder bound, and (3)~the deterministic encoder at $\tau{=}0$ makes the KL formally infinite. The information bottleneck mechanism is preserved---the FM loss increases monotonically with unpredictable content (\Cref{sec:fm_approx})---but AC-UL should be understood as a principled training objective inspired by the ELBO, not as a formal bound.
\end{itemize}

%=====================================================================
\section{Related Work}
\label{sec:related}
%=====================================================================

\textbf{World models for robotics.}
Early work used recurrent state-space models~\citep{ha2018world,hafner2019dream}. DreamerV3~\citep{hafner2023dreamerv3} uses categorical latents. IRIS~\citep{micheli2023transformers} tokenizes observations with a discrete autoencoder. TD-MPC2~\citep{hansen2024tdmpc2} learns latent dynamics without a decoder. DreamZero~\citep{dreamzero2025} introduced autoregressive joint video-action generation at scale.

\textbf{Pretrained features for world models.}
DINO-WM~\citep{zhou2024dinowm} uses frozen DINOv2 patch features. R3M~\citep{nair2022r3m} and VIP~\citep{ma2023vip} learn visual representations for robotics. V-JEPA 2 uses self-supervised video features for zero-shot manipulation.

\textbf{Latent representations for generation.}
Latent Diffusion Models~\citep{rombach2022high} learn autoencoders for diffusion. Unified Latents~\citep{heek2025ul} jointly trains encoders and diffusion priors. Our work extends UL to action-conditioned dynamics.

\textbf{Vision-Language-Action models.}
RT-2~\citep{brohan2023rt2}, OpenVLA~\citep{kim2024openvla}, and GR00T N1~\citep{nvidia2025groot} fine-tune VLMs for action prediction. FLARE~\citep{nvidia2025flare} adds future latent alignment. $\pi_0$~\citep{pi02024} uses flow matching for action generation. Our Stage~2 builds on this paradigm but uses physics-grounded latents.

\textbf{Joint video-action generation.}
DreamZero~\citep{dreamzero2025} and UniSim~\citep{yang2024unisim} jointly generate video and actions. UWM~\citep{zhu2025uwm} couples video and action diffusion. Our work provides a principled latent space for such joint generation.

%=====================================================================
\section{Conclusion}
\label{sec:conclusion}
%=====================================================================

We introduced Action-Conditioned Unified Latents (AC-UL), a framework for adapting pretrained video model latent spaces to encode only the physical geometry predictable from the history of interactions and language instructions.
By applying the Unified Latents variational objective to fine-tune a pretrained video VAE, adding learned action encoding, and training a history-conditioned joint dynamics prior that simultaneously denoises the action latent and the observation latent, we create an information bottleneck that adapts generic video features into action-grounded representations.

Our formulation follows the autoregressive paradigm of DreamZero---using the same Wan-based architecture---but applies the UL variational objective to adapt the latent space for action-relevant content.
Stage~1 fine-tunes the pretrained VAE with the AC-UL objective, and Stage~2 trains an autoregressive diffusion transformer on the frozen adapted latents for joint action-observation generation in closed loop.
By operating on compact, physics-grounded latents, a 1.3B AC-UL model can match or outperform the 14B DreamZero baseline, demonstrating that representation quality outweighs raw model scale.

We believe AC-UL opens promising directions: scaling Stage~2 to foundation-model sizes, incorporating action-free video for broader pretraining, and extending to multi-step chunk-wise generation for real-time control.

%=====================================================================
\bibliographystyle{plainnat}

\begin{thebibliography}{35}

\bibitem[Brohan et~al.(2023)]{brohan2023rt2}
A.~Brohan, N.~Brown, J.~Carbajal, et~al.
\newblock RT-2: Vision-language-action models transfer web knowledge to robotic control.
\newblock In \emph{CoRL}, 2023.

\bibitem[Caron et~al.(2021)]{caron2021dino}
M.~Caron, H.~Touvron, I.~Misra, et~al.
\newblock Emerging properties in self-supervised vision transformers.
\newblock In \emph{ICCV}, 2021.

\bibitem[DreamZero Team(2025)]{dreamzero2025}
DreamZero Team.
\newblock DreamZero: Autoregressive video-action world models for zero-shot robot manipulation.
\newblock \emph{arXiv preprint arXiv:2602.15922}, 2025.

\bibitem[Ha and Schmidhuber(2018)]{ha2018world}
D.~Ha and J.~Schmidhuber.
\newblock World models.
\newblock \emph{arXiv preprint arXiv:1803.10122}, 2018.

\bibitem[Hafner et~al.(2019)]{hafner2019dream}
D.~Hafner, T.~Lillicrap, I.~Fischer, et~al.
\newblock Dream to control: Learning behaviors by latent imagination.
\newblock In \emph{ICLR}, 2019.

\bibitem[Hafner et~al.(2023)]{hafner2023dreamerv3}
D.~Hafner, J.~Pasukonis, J.~Ba, and T.~Lillicrap.
\newblock Mastering diverse domains through world models.
\newblock \emph{arXiv preprint arXiv:2301.04104}, 2023.

\bibitem[Hansen et~al.(2024)]{hansen2024tdmpc2}
N.~Hansen, H.~Su, and X.~Wang.
\newblock TD-MPC2: Scalable, robust world models for continuous control.
\newblock In \emph{ICLR}, 2024.

\bibitem[Heek et~al.(2025)]{heek2025ul}
J.~Heek, E.~Hoogeboom, T.~Mensink, and T.~Salimans.
\newblock Unified Latents (UL): How to train your latents.
\newblock \emph{arXiv preprint arXiv:2602.17270}, 2025.

\bibitem[Ho et~al.(2020)]{ho2020denoising}
J.~Ho, A.~Jain, and P.~Abbeel.
\newblock Denoising diffusion probabilistic models.
\newblock In \emph{NeurIPS}, 2020.

\bibitem[Hoogeboom et~al.(2024)]{hoogeboom2024uvit}
E.~Hoogeboom, J.~Heek, and T.~Salimans.
\newblock Simple diffusion: End-to-end diffusion for high resolution images.
\newblock In \emph{ICML}, 2024.

\bibitem[James et~al.(2020)]{james2020rlbench}
S.~James, Z.~Ma, D.~Rovick~Arrojo, and A.~J.~Davison.
\newblock RLBench: The robot learning benchmark \& learning environment.
\newblock \emph{IEEE RA-L}, 2020.

\bibitem[Khazatsky et~al.(2024)]{khazatsky2024droid}
D.~Khazatsky, K.~Pertsch, S.~Nair, et~al.
\newblock DROID: A large-scale in-the-wild robot manipulation dataset.
\newblock In \emph{RSS}, 2024.

\bibitem[Kim et~al.(2024)]{kim2024openvla}
M.~J.~Kim, K.~Pertsch, S.~Karamcheti, et~al.
\newblock OpenVLA: An open-source vision-language-action model.
\newblock In \emph{CoRL}, 2024.

\bibitem[Kingma et~al.(2021)]{kingma2021variational}
D.~P.~Kingma, T.~Salimans, B.~Poole, and J.~Ho.
\newblock Variational diffusion models.
\newblock In \emph{NeurIPS}, 2021.

\bibitem[Kingma and Gao(2023)]{kingma2023understanding}
D.~P.~Kingma and R.~Gao.
\newblock Understanding diffusion objectives as the ELBO with simple data augmentation.
\newblock In \emph{NeurIPS}, 2023.

\bibitem[Liu et~al.(2024)]{liu2024libero}
B.~Liu, Y.~Zhu, C.~Gao, et~al.
\newblock LIBERO: Benchmarking knowledge transfer for lifelong robot learning.
\newblock In \emph{NeurIPS}, 2024.

\bibitem[Ma et~al.(2023)]{ma2023vip}
Y.~J.~Ma, S.~Sodhani, J.~Jayaraman, et~al.
\newblock VIP: Towards universal visual reward and representation via value-implicit pre-training.
\newblock In \emph{ICLR}, 2023.

\bibitem[Mees et~al.(2022)]{mees2022calvin}
O.~Mees, L.~Hermann, E.~Rosete-Beas, and W.~Burgard.
\newblock CALVIN: A benchmark for language-conditioned policy learning for long-horizon robot manipulation tasks.
\newblock \emph{IEEE RA-L}, 2022.

\bibitem[Micheli et~al.(2023)]{micheli2023transformers}
V.~Micheli, E.~Alonso, and F.~Fleuret.
\newblock Transformers are sample-efficient world learners.
\newblock In \emph{ICLR}, 2023.

\bibitem[Nair et~al.(2022)]{nair2022r3m}
S.~Nair, A.~Rajeswaran, V.~Kumar, et~al.
\newblock R3M: A universal visual representation for robot manipulation.
\newblock In \emph{CoRL}, 2022.

\bibitem[NVIDIA et~al.(2025a)]{nvidia2025flare}
NVIDIA, B.~Wen, S.~Yang, et~al.
\newblock FLARE: Future latent representation alignment for robot policy learning.
\newblock \emph{arXiv preprint arXiv:2505.15659}, 2025.

\bibitem[NVIDIA et~al.(2025b)]{nvidia2025groot}
NVIDIA et~al.
\newblock GR00T N1: An open foundation model for generalist humanoid robots.
\newblock \emph{arXiv preprint}, 2025.

\bibitem[Physical Intelligence(2024)]{pi02024}
Physical Intelligence.
\newblock $\pi_0$: A vision-language-action flow model for general robot control.
\newblock \emph{arXiv preprint}, 2024.

\bibitem[Rombach et~al.(2022)]{rombach2022high}
R.~Rombach, A.~Blatt, D.~Lorenz, P.~Esser, and B.~Ommer.
\newblock High-resolution image synthesis with latent diffusion models.
\newblock In \emph{CVPR}, 2022.

\bibitem[Vahdat et~al.(2021)]{vahdat2021score}
A.~Vahdat, K.~Kreis, and J.~Kautz.
\newblock Score-based generative modeling in latent space.
\newblock In \emph{NeurIPS}, 2021.

\bibitem[Walke et~al.(2023)]{walke2023bridgedata}
H.~R.~Walke, K.~Black, A.~Lee, et~al.
\newblock BridgeData V2: A dataset for robot learning at scale.
\newblock In \emph{CoRL}, 2023.

\bibitem[Yang et~al.(2024)]{yang2024unisim}
M.~Yang, Y.~Du, D.~Hafner, et~al.
\newblock Learning interactive real-world simulators.
\newblock In \emph{ICLR}, 2024.

\bibitem[Zhou et~al.(2024)]{zhou2024dinowm}
G.~Zhou, M.~Liu, et~al.
\newblock DINO-WM: World models on pre-trained visual features enable zero-shot planning.
\newblock \emph{arXiv preprint arXiv:2411.04983}, 2024.

\bibitem[Zhu et~al.(2025)]{zhu2025uwm}
C.~Zhu, Y.~Yu, et~al.
\newblock Unified World Models: Coupling video and action diffusion for pretraining on large robotic datasets.
\newblock \emph{arXiv preprint arXiv:2504.02792}, 2025.

\bibitem[Hu et~al.(2022)]{hu2022lora}
E.~Hu, Y.~Shen, P.~Wallis, et~al.
\newblock LoRA: Low-rank adaptation of large language models.
\newblock In \emph{ICLR}, 2022.

\bibitem[Lipman et~al.(2023)]{lipman2023flow}
Y.~Lipman, R.~T.~Q.~Chen, H.~Ben-Hamu, M.~Nickel.
\newblock Flow matching for generative modeling.
\newblock In \emph{ICLR}, 2023.

\bibitem[Liu et~al.(2023)]{liu2023flow}
X.~Liu, C.~Gong, and Q.~Liu.
\newblock Flow straight and fast: Learning to generate and transfer data with rectified flow.
\newblock In \emph{ICLR}, 2023.

\bibitem[Wan Team(2025)]{wan2025}
Wan Team.
\newblock Wan2.1: Open and advanced large-scale video generative models.
\newblock \emph{arXiv preprint arXiv:2503.20314}, 2025.

\end{thebibliography}

%=====================================================================
\appendix
\section{Detailed Derivations}
\label{app:derivations}
%=====================================================================

\subsection{Full Derivation: From ELBO to Flow Matching Prior}
\label{app:prior_derivation}

We present the complete derivation chain: ELBO $\to$ DDPM KL bound $\to$ flow matching approximation.

\textbf{Step 1: ELBO for joint conditional generation.}
We model the joint distribution $p_\theta(\va_t, \vo_{t+1} \mid \mathcal{H}_t)$, where $\mathcal{H}_t = (\bm{l}, \vo_{t-H+1:t}, \va_{t-H:t-1})$ includes the language instruction. Introducing joint latent variables $(\vz_{t+1,0}, \vb_{t,0})$:
\begin{align}
  -\log p_\theta(\vo_{t+1}, \va_t \mid \mathcal{H}_t)
  &\leq -\E_{q}\!\left[\log p_\theta(\vo_{t+1} | \vz_{t+1,0}) + \log p_\theta(\va_t | \vb_{t,0})\right] \nonumber \\
  &\quad + \KL\!\left[q(\vz_{t+1,0} | \vo_{t+1}) \,\|\, p_\theta(\vz_{t+1,0} | \mathcal{H}_t^z)\right] \nonumber \\
  &\quad + \KL\!\left[q(\vb_{t,0} | \va_t) \,\|\, p_\theta(\vb_{t,0} | \mathcal{H}_t^z)\right].
  \label{eq:app_joint_elbo}
\end{align}

\textbf{Step 2: DDPM KL bound (tight, original UL).}
Following \citet{kingma2021variational}, each KL term is bounded via the diffusion path decomposition. For the observation latent:
\begin{align}
  \KL\!\left[q(\vz_{t+1,0} | \vo_{t+1}) \,\|\, p_\theta(\vz_{t+1,0} | \mathcal{H}_t^z)\right]
  &\leq \E_\tau\!\left[w_{\KL}(\tau)\, \|\vz_{t+1,\text{clean}} - \hat{\vz}_\theta\|^2\right] + C_z,
  \label{eq:app_ddpm_bound}
\end{align}
where $w_{\KL}(\tau) = -\frac{d\lambda(\tau)}{d\tau}\frac{e^{\lambda(\tau)}}{2}$ and $C_z = \KL[q(\vz_{t+1,1} | \vo_{t+1}) \| \mathcal{N}(\bm{0}, \bm{I})]$. An analogous bound holds for the action latent KL.

\textbf{Step 3: Flow matching reweighting.}
For the flow matching interpolation $\vz_\tau = (1-\tau)\vz_{\text{clean}} + \tau\vep$, the KL-optimal weight is $w_{\KL}(\tau) = (1-\tau)/\tau^3$. The flow matching velocity prediction loss (\Cref{prop:fm_equiv}) implicitly uses weight $w_{\text{FM}}(\tau) = 1/\tau^2$. We replace the tight DDPM bound with the flow matching loss:
\begin{align}
  \loss_z^{\text{latent}} &= \E_\tau\!\left[\|v_\theta^z - (\vz_{t+1,\text{clean}} - \vep_z)\|^2\right] = \E_\tau\!\left[\frac{1}{\tau^2}\|\vz_{t+1,\text{clean}} - \hat{\vz}_\theta\|^2\right].
  \label{eq:app_fm_loss}
\end{align}
This is a reweighted version of the DDPM bound: not tight, but preserving the information bottleneck property (the loss increases monotonically with unpredictable content at every $\tau$).

\textbf{Step 4: Action reconstruction and decoder terms.}
The action reconstruction uses a Gaussian likelihood bound: $-\log p_\theta(\va_t | \vb_{t,0}) \leq \|\va_t - D_\theta^a(\vb_{t,\text{clean}})\|^2$.
The decoder term uses the deterministic reconstruction loss $\loss_x$ as an approximate proxy.

\textbf{Step 5: Complete approximate bound.}
\begin{align}
  -\log p_\theta(\vo_{t+1}, \va_t | \mathcal{H}_t) &\;\lessapprox\; \underbrace{\loss_z^{\text{latent}}(\theta)}_{\text{obs.\ prior (FM)}} + \gamma \cdot \underbrace{\loss_b(\theta)}_{\text{action prior (FM)}} + c_{\text{lf}} \cdot \underbrace{\loss_x(\theta)}_{\text{decoder (approx.)}} + \underbrace{\loss_a^{\text{recon}}(\theta)}_{\text{action recon.}}
  \label{eq:app_total}
\end{align}
The approximation stems from three relaxations: (1)~$w_{\text{FM}}$ instead of $w_{\KL}$ for the prior terms, (2)~deterministic decoder instead of diffusion decoder, and (3)~deterministic encoder at $\tau{=}0$ (Dirac delta) makes the prior KL formally infinite, so the ELBO does not hold as a bound. The FM loss instead serves as a practical proxy that preserves the information bottleneck mechanism: it increases monotonically with unpredictable content at every $\tau > 0$.

\subsection{Information-Theoretic Analysis}
\label{app:info_theory}

\textbf{Claim.}
The AC-UL flow matching prior loss provides a reweighted bound on $I(\vz_{t+1,\text{clean}}; \vo_{t+1} \mid \mathcal{H}_t^z)$.

\textbf{Proof.}
The KL in the ELBO satisfies:
\begin{align}
  \E_{p(\mathcal{H}_t)}\!\left[\KL\!\left[q(\vz_{t+1,0} | \vo_{t+1}) \,\|\, p_\theta(\vz_{t+1,0} | \mathcal{H}_t^z)\right]\right]
  &\geq I(\vz_{t+1,0}; \vo_{t+1} | \mathcal{H}_t^z).
\end{align}
When the prior is optimal ($p_\theta = p^*$), equality holds. The DDPM KL bound with $w_{\KL}$ upper-bounds the KL; the flow matching loss with $w_{\text{FM}}$ is a reweighted version. Since both weightings are strictly positive for all $\tau \in (0,1)$, $\loss_z^{\text{latent}}$ increases monotonically with $I(\vz_{t+1,\text{clean}}; \vo_{t+1} | \mathcal{H}_t^z)$, preserving the information bottleneck.

\textbf{Role of action latent loss.}
The action latent loss $\loss_b$ provides an additional constraint via the same flow matching framework. The reweighted KL bound gives:
\begin{equation}
  \loss_b \geq I(\vb_{t,\text{clean}}; \va_t \mid \mathcal{H}_t^z).
\end{equation}
At optimality, the joint denoising of $(\vb_t, \vz_{t+1})$ ensures that the action latent and observation latent are mutually informative:
\begin{equation}
  I(\vb_{t,\text{clean}}; \vz_{t+1,\text{clean}} | \mathcal{H}_t^z) \geq I(\va_t; \vo_{t+1} | \mathcal{H}_t^z) - I(\va_t; \vo_{t+1} | \vz_{t+1,\text{clean}}, \mathcal{H}_t^z).
\end{equation}
The joint flow matching loss pushes $\vz_{t+1,\text{clean}}$ to be a \emph{sufficient statistic} for the action given the history, meaning $I(\va_t; \vo_{t+1} | \vz_{t+1,\text{clean}}, \mathcal{H}_t^z) \to 0$. This ensures the observation latent captures the full causal effect of the action, while the action encoder learns a representation that is jointly denoised with the observation latent.

%=====================================================================
\section{Implementation Details}
\label{app:implementation}
%=====================================================================

\subsection{Stage 1 Hyperparameters}

\begin{table}[h]
\centering
\caption{Stage 1 hyperparameters.}
\label{tab:app_stage1}
\begin{tabular}{ll}
\toprule
\textbf{Hyperparameter} & \textbf{Value} \\
\midrule
Image resolution & $320 \times 176$ (multi-view composite) \\
Latent channels & 16 (Wan2.1 VAE) \\
Spatial downsampling & $8\times$ (Wan2.1 VAE) \\
VAE initialization & Pretrained Wan2.1 \\
VAE fine-tuning & LoRA (rank 4, alpha 4) \\
Prior initialization & Pretrained Wan DiT (1.3B) \\
Prior fine-tuning & LoRA (rank 4) \\
History length $H$ & 4 \\
Noise schedule & Flow matching (linear interpolation) \\
Timestep sampling $p(\tau)$ & Shifted Beta (shift $= 5.0$) \\
Loss factor $c_{\text{lf}}$ & 1.5 \\
Action latent loss weight $\gamma$ & 1.0 \\
Action encoder $A_\theta$ & Embodiment-specific MLP \\
Action latent dim $d_b$ & 64 \\
Action decoder $D_\theta^a$ & Embodiment-specific MLP \\
State encoder & Embodiment-specific MLP \\
Text encoder & UMT5-XXL (frozen) \\
Optimizer & AdamW ($\beta_1 = 0.95$, $\beta_2 = 0.999$) \\
Learning rate & $1 \times 10^{-4}$ \\
LR schedule & Cosine with 5K warmup \\
Batch size & 8 (per GPU) $\times$ 8 GPUs \\
Training steps & 500K \\
Precision & bf16 \\
DeepSpeed & ZeRO Stage 2 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Stage 2 Hyperparameters}

\begin{table}[h]
\centering
\caption{Stage 2 hyperparameters.}
\label{tab:app_stage2}
\begin{tabular}{ll}
\toprule
\textbf{Hyperparameter} & \textbf{Value} \\
\midrule
Architecture & CausalWanModel (1.3B) \\
Initialization & Pretrained Wan DiT \\
Fine-tuning & LoRA (rank 4) \\
Frames per block & 2 \\
Actions per block & 24 \\
States per block & 1 \\
Text encoder & UMT5-XXL (frozen) \\
Image encoder & OpenCLIP ViT-H/14 (frozen) \\
Training objective & Flow matching \\
Noise schedule & Beta(1.5, 1.0) \\
Denoising steps (inference) & 4 \\
Optimizer & AdamW ($\beta_1 = 0.95$, $\beta_2 = 0.999$) \\
Learning rate & $1 \times 10^{-4}$ \\
LR schedule & Cosine with 2K warmup \\
Batch size & 8 (per GPU) $\times$ 8 GPUs \\
Precision & bf16 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Planning Hyperparameters}

\begin{table}[h]
\centering
\caption{Planning hyperparameters.}
\label{tab:app_planning}
\begin{tabular}{ll}
\toprule
\textbf{Hyperparameter} & \textbf{Value} \\
\midrule
Best-of-$N$ samples & 256 \\
Planning horizon $P$ & 10 steps \\
Receding horizon (execute) & $K$ = 2 steps \\
Action smoothing & Savitzky-Golay filter \\
\bottomrule
\end{tabular}
\end{table}

\end{document}
